--- 
title: "`sangchem` R package: research project data analysis documentation"
author: "Tomasz Włodarczyk"
date: "2023-12-10"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
description: |
  This is a doucumentation of the data analysis workflows and function used to generate the results included in the manuscript entitled Coming out from the shadows: facultative slave-making ants reveal their chemical identity during colony development. 
link-citations: yes
github-repo: TomVuod/sangchem
---
# Introduction {-}
Here we present how the result shown in the manuscript were obtained starting from the data sets collected during the study. The dedicated [`sangchem`](https://github.com/TomVuod/sangchem) R package  is available on GitHub and can be isstalled via:
```{r eval=FALSE}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("TomVuod/sangchem")
```

<!--chapter:end:index.Rmd-->

# Changes of CHCs amount over time

In this report we will check how the all CHCs or their specific subset changes in ants
of both species as colony develops. 

```{r}
suppressMessages({
library(sangchem)
library(dplyr)
library(lme4)
library(lmerTest)
library(DHARMa)
})
# load data
data("development_data")
load_globals()
dev_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==0) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(corrected_mass=mass/head_width^2*10^6)
```

## Change in the total CHCs mass
### Mature *F. sanguinea*
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop  + (1|colony) +(1 | colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

### *F. fusca*
```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop  + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

### Callow *F. sanguinea*
```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(corrected_mass=mass/head_width^2*10^6)

lm_model <- lmer(I((corrected_mass)) ~ sang_prop  + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

## Change of the CHCs associated with *F. sanguinea*
### *F. sanguinea* ants
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop +(1 | colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

### *F. fusca* ants
```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lmer(corrected_mass ~ sang_prop +(1|colony:census_date) + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

## Change of the CHCs associated with *F. fusca*
### *F. sanguinea* ants
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop +(1 | colony:census_date) + (1 | colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

### *F. fusca* ants
```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop + (1 | colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

<!--chapter:end:CHCs_over_time.Rmd-->

# Species discrimination and prediction

## Species markers
This  report shows how the peaks discriminative for *F. sanguinea* and *F. fusca* were determined. We will use the chemical data collected from the filed colonies in the earlier study[@WLODARCZYK201798]. In this way our method of marker selection is blind to the dataset to which  markers will be applied avoiding potential source of the bias in the analysis results.
```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(stringi)
  library(dplyr)
  library(mixOmics)
})

# load data from the the earlier study
data("field_colonies")

# select samples from F. sanguinea ants and free-living F. fusca ants encoded in chromatogram_ID
fusca_samples <- as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID), "fus[0-9].?-[0-9]\\([1-2]\\)")])
sanguinea_samples<-as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID),"sangFs[0-9].?\\([0-9]\\)")])

# remove outliers
sanguinea_samples <- setdiff(sanguinea_samples,c("sangFs2(1)","sangFs5(2)","sangFs26(1)","sangFs26(2)"))
fusca_samples <- setdiff(fusca_samples,c("fus9-2(1)", "fus26-3(2)"))
discrimination_data <- filter(field_colonies,chromatogram_ID %in% c(fusca_samples,sanguinea_samples)) 
peak_prop_unfiltered <- peak_proportions_table(discrimination_data)
# feature selection
peak_prop <- freq_abun_QC(peak_prop_unfiltered)
# re-normalize rows to 1
peak_prop <- peak_prop/rowSums(peak_prop)
# replace zeros with a small number before transformation
peak_prop[peak_prop==0] <- 10^-16
# apply central log ratio transformation
peak_transformed <- t(apply(peak_prop, 1, clr_transformation))
```

```{r}
PCA_species_discr <- prcomp(peak_transformed, scale. = TRUE)
```

We also need to deifne our reponse variable to be explained by discrimination model.
```{r}
#define groups to be compared
Y <- rownames(PCA_species_discr$x) %in% sanguinea_samples
```

Now that we have prepared the input data we can run the model. The following workflow is based on the tutorial available at the official [`mixOmics` package web site](https://mixomics.org/mixmc/hmp-bodysites-case-study/).
```{r}
# run the main model
species_discrimination_PC.plsda <- plsda(PCA_species_discr$x, Y, ncomp=10, scale=FALSE)
plotIndiv(species_discrimination_PC.plsda,group=Y,
          legend=TRUE,ellipse=TRUE,comp=1:2)

# determine the optimal number of components used in the model
spec_discr_PC.plsda.perf<-perf(species_discrimination_PC.plsda, nrepeat = 10,folds=5,
                            progressBar = FALSE,validation = "Mfold",
                            auc=TRUE, scale=FALSE)

plot(spec_discr_PC.plsda.perf, sd = TRUE, overlay = 'measure')
```

The diagnstic plots suggests that we should keep 2 components for the final model.
```{r}
selected_n <- 2

# dtermine the number of the original variables to be kept in each component (actually
# here we have only one component)
spec_discr_PC.splsda <- tune.splsda(PCA_species_discr$x,Y,test.keepX = c(1:60),ncomp = selected_n,
                                 validation="Mfold", progressBar=FALSE, dist="max.dist", 
                                 nrepeat=5, cpus=2, folds=5, measure="BER", scale=FALSE)

# final model
species_prediction_model <- splsda(PCA_species_discr$x, Y, ncomp=selected_n,
                           keepX =spec_discr_PC.splsda$choice.keepX, scale=FALSE)

# visuzalize results for the first two commponents
plotIndiv(species_prediction_model, group=Y, legend=TRUE, ellipse=TRUE, comp=1:2)

```

Now we want to correlate relative abundances of CHCs with the species predicted by the model for each sample.
```{r message=FALSE}
# quantify predicted species (F. sanguinea closer to one and F. fusca closer to 0)
predicted_species <- predict(species_prediction_model, PCA_species_discr$x)$predict[,2,species_prediction_model$ncomp]

library(Hmisc) 
cor_p_val<-rcorr(cbind(peak_prop, predicted_species))
cor_p_val$P[cor_p_val$r==1]<-0
#significance threshold = 0.05
significant_correlations <- cor_p_val$r["predicted_species",][cor_p_val$P["predicted_species",]<0.05]
peaks_sang <- names(significant_correlations)[significant_correlations>0]
peaks_sang <- as.numeric((setdiff(peaks_sang,"predicted_species")))
peaks_fusca <- as.numeric(names(significant_correlations)[significant_correlations<0])
print(peaks_sang)
print(peaks_fusca)
```

The objects `peaks_sang`, `peaks_fusca`, `species_prediction_model` and `PCA_species_discr` are stored in the package and can be collectively loaded into the global environment along with the other objects by running `load_globals()`.
```{r eval=FALSE}
globals <- readRDS(system.file("globals.rds", package="sangchem"))
globals[["species_prediction_model"]] <- species_prediction_model
globals[["PCA_species_discr"]] <- PCA_species_discr
globals[["peak_sang"]] <- peaks_sang
globals[["peak_fusca"]] <- peaks_fusca
saveRDS(globals, system.file("globals.rds", package="sangchem"))
```

### Peak position on the species discriminant axis

We might ask how a given peak compares to other in term of being characteristic more of one or the other species. Correspondingly we will generate the artificial profiles which are composed of only one peak to see how each peak contributes to the species prediction.

```{r}
unit_input <- diag(rep(1, ncol(peak_prop)))
colnames(unit_input) <- rownames(unit_input) <-colnames(peak_prop)
res <- calculate_SII(unit_input)
colnames(res)[1] <- "peak ID"
# normalize scores
res[,2] <- res[,2] - mean(res[,2])
res[,2] <- res[,2]/sd(res[,2])
print(res[order(res$predicted_species),])
```

## Species identity score
Ants in mixed colonies mixed CHCs to produce hybrid chemical identity. Here we develop the method to assign each sample a number indicating how much it is characteristic of one or the other species.
The main idea is to fit a sPLS_DA model to the training set (sampled from field colonies) and use it predict the explanatory variable (species) in the test data (samples from experimental colonies). However the sPLS_DA model run on the raw data will yield the classificator which used only few variables (peaks) due to the well separation of samples of both species. The unused variables might also be informative in the species discrimination task. We want to use as much data as it is possible to make robust predictions and thus we will use principal component analysis output. Thus, the rotation matrix from the principal component analysis will be used to transform input data to the discrimination model. The correspondig code in encapsulated in the `claculate_SII` function.

### Data preparation
```{r}
data("development_data")
dev_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony")) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, callow, census_date)
normalized_abundances <- peak_proportions_table(mass_spectra_data[mass_spectra_data$chromatogram_ID %in% dev_samples$chromatogram_ID,])
species_indices <- calculate_SII(normalized_abundances)
dev_samples <- left_join(dev_samples, species_indices)
```

### Predicted species of different categories of ants as a function of the proportion of *F. sanguinea* ants in a colony
#### Mature *F. sanguinea*
```{r}
library(lmerTest)
library(DHARMa)
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==0,]
lm_model <- lmer(predicted_species~sang_prop+(1|colony)+(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

#### *F. fusca*
```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
lm_model <- lmer(predicted_species~sang_prop+(1|colony)+(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

#### Callow *F. sanguinea*
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==1,]
lm_model <- lmer(predicted_species~sang_prop+(1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

### Difference in species prediction between mature *F. sanguinea* ants and their slaves
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==0,]
model_input$index_diff <- model_input$predicted_species - find_nestmates_SII(model_input$chromatogram_ID, dev_samples)
lm_model <- lmer(index_diff~sang_prop+(1|colony)+(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

### Difference in species prediction between callow *F. sanguinea* ants and their slaves

Since the linear model does not meet one of diagnostics criterion we will apply Wilcoxon paired test.
```{r message=FALSE}
library(ggpubr)
library(rstatix)
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==1,]
model_input$slave_species_index <- find_nestmates_SII(model_input$chromatogram_ID, dev_samples)
print(wilcox.test(model_input$predicted_species, model_input$slave_species_index, paired=TRUE))
df <- data.frame(species_identity_index=c(model_input$predicted_species, model_input$slave_species_index), species="F. fusca")
df$species[1:nrow(model_input)] <- "callow F. sanguinea"
stat_tests <- wilcox_test(df, species_identity_index~species, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="species", dodge=0.8, fun="max")
ggboxplot(df, x="species", y="species_identity_index")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(stat_tests, label="p", tip.length = 0, size=3.5)
```

### Difference in species prediction between callow and mature *F. sanguinea* ants.
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==1,]
model_input$index_diff <- model_input$predicted_species - find_nestmates_SII(model_input$chromatogram_ID, dev_samples, heterospecific = FALSE)
# no random effect since it captures no variance
lm_model <- lm(index_diff~sang_prop, data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
```

## Session info
```{r}
print(sessionInfo())
```

<!--chapter:end:Species_identity.Rmd-->

# Finding peaks discriminating callow and mature *F. sanguinea* workers

```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(dplyr)
  library(mixOmics)
})

# load data
data("development_data")
data("mass_spectra_data")

DA_data <- dplyr::filter(development_data, caste=="worker", !remarks %in% c("aggression actor", "aggression target", "queenless colony")) %>%
  dplyr::select(colony, species, callow, census_date, chromatogram_ID, sang_prop) %>%
  split(list(.$colony, .$census_date), drop=TRUE) %>%
  lapply(function(x){if(sum(x$callow)==0) return(NULL); x}) %>%
  {function(x) x[!unlist((lapply(x, is.null)))]}() %>%
  purrr::map2(seq_along(.), function(x,y){x$group <- y; x}) %>%
  {function(x) do.call(rbind, x)}()

Y <- (DA_data$callow==1) + (DA_data$species=="F. sanguinea")

peak_prop <- peak_proportions_table(mass_spectra_data[mass_spectra_data$chromatogram_ID %in% DA_data$chromatogram_ID,])
peak_prop <- peak_prop[match(DA_data$chromatogram_ID, rownames(peak_prop)),]
# replace zeros with a small number before transformation
peak_prop[peak_prop==0] <- 10^-16
# apply central log ratio transformation
peak_transformed <- t(apply(peak_prop, 1, clr_transformation))
PCA_callow_discr <- prcomp(peak_transformed, scale. = TRUE)
```
```{r}
discr_analysis_callow <- splsda(PCA_callow_discr$x, Y=Y ,multilevel=DA_data$group, 
                                keepX=seq(1:60), ncomp=7, scale = FALSE)
plotIndiv(discr_analysis_callow, legend=TRUE,ellipse=TRUE,comp=1:2)

perf.discr_analysis_callow <- perf(discr_analysis_callow, 
                                validation = "Mfold", 
                                folds = 4, 
                           progressBar = FALSE, 
                           auc = TRUE, 
                           nrepeat = 10,
                           scale = FALSE)
plot(perf.discr_analysis_callow, col = color.mixo(1:3), sd = TRUE, legend.position = "horizontal")
selected_n <- 5
discr_analysis_callows_tuned <- tune.splsda(PCA_callow_discr$x,
                                   Y=Y,
                                   multilevel=DA_data$group,
                                   ncomp = selected_n,
                                   test.keepX = seq(1:120), 
                                   dist = 'mahalanobis.dist',
                                   validation = 'loo',
                                   progressBar = FALSE,
                                   scale = FALSE) 

discr_analysis_callows_res <- splsda(PCA_callow_discr$x,Y=Y,
                      multilevel= DA_data$group,
                      keepX=discr_analysis_callows_tuned$choice.keepX,
                      ncomp=selected_n,
                      scale = FALSE)

plotIndiv(discr_analysis_callows_res,
          legend=TRUE, ellipse=TRUE, comp=1:2)
```

## Determining callow markers
```{r}
# quantify predicted species (callow workers closer to one)
predicted_species <- calculate_SII(peak_prop, discr_analysis_callows_res, PCA_callow_discr, predicted_category=3)
# build matrix to calculate correlation of `prediction score` and peak relative abundances 
predicted_species <- matrix(predicted_species$predicted_species[match(rownames(peak_prop), predicted_species$chromatogram_ID)], ncol=1, dimnames = list(rownames(peak_prop), "predicted_species"))
library(Hmisc) 
cor_p_val<-rcorr(cbind(peak_prop, predicted_species))
cor_p_val$P[cor_p_val$r==1]<-0
# significance threshold = 0.05
significant_correlations <- cor_p_val$r["predicted_species",][cor_p_val$P["predicted_species",]<0.05]
peaks_callow_sig <- names(significant_correlations)[significant_correlations>0]
peaks_callow_sig <- as.numeric((setdiff(peaks_callow_sig,"predicted_species")))

# checking the effect of each peak on the prediction score for each of the
# three categories (F. fusca, mature F. sanguinea, callow F. sanguinea)

# prepare input data (one peak present per one sample)
unit_input <- diag(rep(1,ncol(peak_prop)))
colnames(unit_input) <- rownames(unit_input) <-colnames(peak_prop)
# predict each category for each peak
peak_prediction <- purrr::reduce(lapply(1:3, function(x) calculate_SII(unit_input, discr_analysis_callows_res, PCA_callow_discr, predicted_category=x)), merge, by="chromatogram_ID")
colnames(peak_prediction)[1] <- "peak ID"
peak_prediction[,2:4] <- apply(peak_prediction[,2:4], 2, function(x) {x <- x-mean(x); x/sd(x)})
colnames(peak_prediction)[2:4] <- c("fusca", "sanguinea", "callow") 
# choose peaks positively associated with being callow and negatively with two other categories
peaks_callow_predictors <- peak_prediction[peak_prediction$fusca<0&peak_prediction$sanguinea<0&peak_prediction$callow>0,"peak ID"]
peaks_callow <- intersect(peaks_callow_sig, peaks_callow_predictors)
print(peaks_callow)

# peaks negatively related with being a callow worker
top_negative <- peak_prediction[order(peak_prediction$callow),][1:10,c("peak ID", "callow")]
top_negative[["sanguinea marker"]] <- top_negative[["peak ID"]] %in% peaks_sang
top_negative[["fusca marker"]] <- top_negative[["peak ID"]] %in% peaks_fusca
rownames(top_negative) <- NULL
print(top_negative)
```

## Determining mature *F. sanguinea* markers
```{r}
# quantify predicted species (mature F. sanguinea closer to one)
predicted_species <- calculate_SII(peak_prop, discr_analysis_callows_res, PCA_callow_discr, predicted_category=2)
# build matrix to calculate correlation of `prediction score` and peak relative abundances 
predicted_species <- matrix(predicted_species$predicted_species[match(rownames(peak_prop), predicted_species$chromatogram_ID)], ncol=1, dimnames = list(rownames(peak_prop), "predicted_species"))
library(Hmisc) 
cor_p_val<-rcorr(cbind(peak_prop, predicted_species))
cor_p_val$P[cor_p_val$r==1]<-0
# significance threshold = 0.05
significant_correlations <- cor_p_val$r["predicted_species",][cor_p_val$P["predicted_species",]<0.05]
peaks_mature_sig <- names(significant_correlations)[significant_correlations>0]
peaks_mature_sig <- as.numeric((setdiff(peaks_mature_sig, "predicted_species")))
# choose peaks positively associated with being mature F. sanguinea and negatively with two other categories
peaks_mature_predictors <- peak_prediction[peak_prediction$fusca<0&peak_prediction$sanguinea>0&peak_prediction$callow<0,"peak ID"]
peaks_mature <- intersect(peaks_mature_sig, peaks_mature_predictors)
print(peaks_mature)
```

```{r eval=FALSE}
globals <- readRDS(system.file("globals.rds", package="sangchem"))
globals[["peaks_callow"]] <- peaks_callow
globals[["peaks_mature"]] <- peaks_mature
saveRDS(globals, system.file("globals.rds", package="sangchem"))
```

<!--chapter:end:Callow_markers.Rmd-->

# Change in dissimilarity between callow and mature *F. sanguinea* as the proportion of slave-makers in the colony grows

```{r}

library(sangchem)
library(dplyr)
library(lme4)
library(lmerTest)
library(DHARMa)
data("mass_spectra_data")
data("development_data")
development_data_clean <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), species=="F. sanguinea") %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date, callow, caste)
get_distance_to_mature <- function(callow_chr_ID, mature_chr_IDs, MS_data){
  mean(unlist(lapply(mature_chr_IDs, pair_distance, callow_chr_ID, MS_data)))

}
callow_data <- filter(development_data_clean, callow==1) 
mature_chr_IDs <- lapply(callow_data$chromatogram_ID, find_sample_nestmate, heterospecific=FALSE)
dissimilarities <- mapply(get_distance_to_mature, callow_data$chromatogram_ID, mature_chr_IDs, MoreArgs= list(MS_data=mass_spectra_data))
model_input <- cbind(data.frame(dissimilarity=dissimilarities), callow_data[,c("colony", "census_date", "sang_prop")])
lm_model <- lmer(dissimilarity~sang_prop + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

<!--chapter:end:Callow-mature_dissimilarity.Rmd-->

