# Body surface area of *F. sanguinea* ants and their slaves {#surface-area}

The areas of planar projection of different body parts were regressed on the head width to get the model which was then used to approximate area for each sample. The following formula has been applied

$$
\hat{A} = \sum_{i \in I}\beta_{0,i}+\beta_{1,i}w+\beta_{2,i}x^2,
$$
where
$\hat{A}$ denotes estimated body area
$I$ denotes indices of the body parts for which area of the planar projection was measured
$w$ denotes head width
$\beta_{j,i}$ is a coefficient from the linear regression 

These calculations do not reflect the absolute value of body surface area but are transformed into scaling factors by taking their ratio to a standard area (estimated on the same scale). In this way, the unit of measurement is canceled out, and absolute values are converted into dimensionless ratios.

```{r warning=FALSE}
library(sangchem)
library(ggplot2)
data("development_data")
data("separation_experiment_data")
data("body_surface_area")

body_surface_area$species <- development_data$species[match(body_surface_area$ID, development_data$chromatogram_ID)]
body_surface_area$head_width <- development_data$head_width[match(body_surface_area$ID, development_data$chromatogram_ID)]

```
```{r fig.cap="Relation between head width and area of the planar projections of ant body parts.",fig.height=3}
ggplot(body_surface_area, aes(y=area, x=head_width,color=species, shape = view))+
  geom_point()+
  stat_smooth(method='lm', formula=y~poly(x,2,raw = TRUE), se=FALSE) +
  facet_wrap(~view)+
  scale_color_manual(values=c("black", "red"))+
  xlab(expression(paste("Head width [",mu,"m]")))+
  ylab("Area of the planar projection [pixels]")+
  theme_light()
```
