# Finding peaks discriminating callow and mature *F. sanguinea* workers

The procedure applied to identify species markers can also be applied to determine peaks characteristic of *F. sanguinea* callow ants. In discriminant analysis, we include two more groups: adult *F. sanguinea* and adult *F. fusca* as contrasts against which the callow features are extracted. 

```{r warning=FALSE}
# devtools::install_github("mixOmicsTeam/mixOmics", ref = "143d31f9629bc83f1f27272ba6b7a2aaef0ee0e8")
suppressMessages({
  library(sangchem)
  library(dplyr)
  library(mixOmics)
})

# load data
data("development_data")
data("mass_spectra_data")

DA_data <- dplyr::filter(development_data, caste=="worker", !remarks %in% c("aggression actor", "aggression target", "queenless colony")) %>%
  dplyr::select(colony, species, callow, census_date, chromatogram_ID, sang_prop) %>%
  split(list(.$colony, .$census_date), drop=TRUE) %>%
  lapply(function(x){if(sum(x$callow)==0) return(NULL); x}) %>%
  {function(x) x[!unlist((lapply(x, is.null)))]}() %>%
  purrr::map2(seq_along(.), function(x,y){x$group <- y; x}) %>%
  {function(x) do.call(rbind, x)}()

Y <- (DA_data$callow==1) + (DA_data$species=="F. sanguinea")

peak_prop <- peak_proportions_table(mass_spectra_data[mass_spectra_data$chromatogram_ID %in% DA_data$chromatogram_ID,])
peak_prop <- peak_prop[match(DA_data$chromatogram_ID, rownames(peak_prop)),]
# remove absent peaks
peak_prop <- peak_prop[,apply(peak_prop,2,function(x) sum(x)>0)]
peak_prop <- freq_abun_QC(peak_prop, freq_cutoff = 0.3, abundance_cutoff = 0.005)
# replace zeros with a small number before transformation
peak_prop[peak_prop==0] <- 10^-16
# apply central log ratio transformation
peak_transformed <- t(apply(peak_prop, 1, clr_transformation))
PCA_callow_discr <- prcomp(peak_transformed, scale. = TRUE)
# PCA_callow_discr$x <- withinVariation(PCA_callow_discr$x, data.frame(DA_data$group))
```
```{r}
set.seed(1342)
discr_analysis_callow <- splsda(PCA_callow_discr$x, Y=Y ,multilevel=DA_data$group, 
                                ncomp=7, scale = FALSE)
plotIndiv(discr_analysis_callow, legend=TRUE,ellipse=TRUE,comp=1:2)

perf.discr_analysis_callow <- perf(discr_analysis_callow, 
                                validation = "Mfold", 
                                folds = 4, 
                           progressBar = FALSE, 
                           auc = TRUE, 
                           nrepeat = 30,
                           scale = FALSE)
plot(perf.discr_analysis_callow, col = color.mixo(1:3), sd = TRUE, legend.position = "horizontal")
selected_n <- perf.discr_analysis_callow$choice.ncom["overall", "max.dist"]
discr_analysis_callows_tuned <- tune.splsda(PCA_callow_discr$x,
                                   Y=Y,
                                   nrepeat = 30,
                                   folds = 4,
                                   ncomp = selected_n,
                                   test.keepX = seq(1:dim(PCA_callow_discr$x)[2]), 
                                   dist = 'max.dist',
                                   validation = 'Mfold',
                                   measure = "BER",
                                   progressBar = FALSE,
                                   scale = FALSE) 

selected_n <- discr_analysis_callows_tuned$choice.ncomp$ncomp
discr_analysis_callows_res <- splsda(PCA_callow_discr$x,Y=Y,
                      multilevel= DA_data$group,
                      keepX=discr_analysis_callows_tuned$choice.keepX,
                      ncomp=selected_n,
                      scale = FALSE)

plotIndiv(discr_analysis_callows_res,
          legend=TRUE, ellipse=TRUE, comp=1:2)
```

Now we are ready to determine the marker peaks. To do so, we will calculate the effect of each peak on the final classification. We generate input matrix populated with all ones and then we increment input for one peak at a time. 

## Determining callow markers
```{r}
ones_matrix <- matrix(1, nrow = ncol(peak_prop),ncol= ncol(peak_prop))
colnames(ones_matrix) <- rownames(ones_matrix) <- colnames(peak_prop)
unit_input <- diag(rep(1, ncol(peak_prop)))
incremented_matrix <- ones_matrix + unit_input
ones_matrix <- ones_matrix[1,,drop=FALSE]
ones_matrix <- ones_matrix/ncol(ones_matrix)
incremented_matrix <- apply(incremented_matrix, 2, function(x) x/sum(x))
res1 <- calculate_SII(incremented_matrix, discr_analysis_callows_res, PCA_callow_discr, predicted_category=3)
res2 <- calculate_SII(ones_matrix, discr_analysis_callows_res, PCA_callow_discr, predicted_category=3)
res <- res1
res$predicted_callow = res1$predicted_species - res2$predicted_species
res <- res[order(res$predicted_callow),]
colnames(res)[1] <- "peak_ID"
compounds <- c()
for(peak in res$peak_ID){
  compounds <- c(compounds, mass_spectra_data[mass_spectra_data$peak_ID==peak,"identification"][1])
}
res$compounds = compounds
res <- res[,c("peak_ID", "compounds", "predicted_callow")]
print(res)
peaks_callow <- res$chromatogram_ID[res$predicted_callow<=quantile(res$predicted_callow, 0.2)]
```

## Determining mature *F. sanguinea* markers
```{r}
res1 <- calculate_SII(incremented_matrix, discr_analysis_callows_res, PCA_callow_discr, predicted_category=2)
res2 <- calculate_SII(ones_matrix, discr_analysis_callows_res, PCA_callow_discr, predicted_category=2)
res <- res1
res$predicted_mature = res1$predicted_species - res2$predicted_species
res <- res[order(res$predicted_mature),]
colnames(res)[1] <- "peak_ID"
compounds <- c()
for(peak in res$peak_ID){
  compounds <- c(compounds, mass_spectra_data[mass_spectra_data$peak_ID==peak,"identification"][1])
}
res$compounds = compounds
res <- res[,c("peak_ID", "compounds", "predicted_mature")]
print(res)
peaks_mature <- res$chromatogram_ID[res$predicted_mature<=quantile(res$predicted_mature, 0.2)]
```

```{r eval=params$save_globals}
globals <- readRDS(system.file("globals.rds", package="sangchem"))
globals[["discr_analysis_callows_res"]] <- discr_analysis_callows_res
globals[["PCA_callow_discr"]] <- PCA_callow_discr
globals[["peaks_mature"]] <- as.numeric(peaks_mature)
globals[["peaks_mature"]] <- as.numeric(peaks_mature)
saveRDS(globals, system.file("globals.rds", package="sangchem"))
```

## Verifying the performace of the discrimination model

We will verify the performance the above model against the null ones which are trained using the data where age category of *F. sanguinea* ants has been shuffled. 

```{r eval = FALSE}
# running basic model 
set.seed(1342)
results_basic <- c()
while(length(results_basic)<1000){
  message(counter)
  .Random.seed <- results$random.seed
  discr_analysis_callow <- splsda(PCA_callow_discr$x, Y=Y ,
                                  ncomp=7, scale = FALSE)
  tryCatch({perf.discr_analysis_callow <- perf(discr_analysis_callow,
                                               validation = "Mfold",
                                               folds = 4,
                                               progressBar = FALSE,
                                               auc = TRUE,
                                               nrepeat = 30,
                                               scale = FALSE,
                                               cpus=16)
  
  selected_n <- perf.discr_analysis_callow$choice.ncom["overall", "max.dist"]
  discr_analysis_callows_tuned <- tune.splsda(PCA_callow_discr$x,
                                              Y=Y,
                                              ncomp = selected_n,
                                              test.keepX = seq(1:120),
                                              validation = 'Mfold',
                                              folds = 4, 
                                              nrepeat = 30,
                                              dist = 'max.dist', # use max.dist measure
                                              measure = "BER",
                                              progressBar = FALSE,
                                              scale = FALSE)
  
  n_comp <- discr_analysis_callows_tuned$choice.ncomp$ncomp
  if(is.null(n_comp)) n_comp <- 1
  discr_analysis_callows_res <- splsda(PCA_callow_discr$x,Y=Y,
                                       keepX=discr_analysis_callows_tuned$choice.keepX,
                                       ncomp=n_comp,
                                       scale = FALSE)
  
  perf.discr_analysis_callows_res <- perf(discr_analysis_callows_res,
                                          validation = "Mfold",
                                          folds = 4,
                                          progressBar = FALSE,
                                          nrepeat = 1,
                                          scale = FALSE,
                                          cpus=16)
  
  curve_data <- calculate_accuracy_metrics(perf.discr_analysis_callows_res$predict[[length(perf.discr_analysis_callows_res$predict)]][,3,1][Y>0], Y[Y>0]==2)
  AUC <- calculate_AUC(curve_data$FPR, curve_data$TPR)
  results_basic <- c(results_basic, AUC)}, 
  error=function(cond) message(cond))
}


# running null models
set.seed(1342)
results_null <- c()
while(length(results_null)<1000){
  # shuffle categories 1 (mature F. sanguinea) and 2 (callow F. sanguinea)
  Y <- split(Y, DA_data$group) %>% lapply(function(x) {x[x>0]<-sample(x[x>0], length(x[x>0]));x}) %>% unlist()
  discr_analysis_callow <- splsda(PCA_callow_discr$x, Y=Y ,
                                  ncomp=7, scale = FALSE)
  tryCatch({perf.discr_analysis_callow <- perf(discr_analysis_callow,
                                               validation = "Mfold",
                                               folds = 4,
                                               progressBar = FALSE,
                                               auc = TRUE,
                                               nrepeat = 30,
                                               scale = FALSE,
                                               cpus=16)
  
  selected_n <- perf.discr_analysis_callow$choice.ncom["overall", "max.dist"]
  discr_analysis_callows_tuned <- tune.splsda(PCA_callow_discr$x,
                                              Y=Y,
                                              ncomp = selected_n,
                                              test.keepX = seq(1:120),
                                              validation = 'Mfold',
                                              folds = 4, 
                                              nrepeat = 30,
                                              dist = 'max.dist', # use max.dist measure
                                              measure = "BER",
                                              progressBar = FALSE,
                                              scale = FALSE,
                                              cpus=16)
  
  n_comp <- discr_analysis_callows_tuned$choice.ncomp$ncomp
  if(is.null(n_comp)) n_comp <- 1
  discr_analysis_callows_res <- splsda(PCA_callow_discr$x,Y=Y,
                                       keepX=discr_analysis_callows_tuned$choice.keepX,
                                       ncomp=n_comp,
                                       scale = FALSE)
  
  perf.discr_analysis_callows_res <- perf(discr_analysis_callows_res,
                                          validation = "Mfold",
                                          folds = 4,
                                          progressBar = FALSE,
                                          nrepeat = 1,
                                          scale = FALSE,
                                          cpus=16)
  
  curve_data <- calculate_accuracy_metrics(perf.discr_analysis_callows_res$predict[[length(perf.discr_analysis_callows_res$predict)]][,3,1][Y>0], Y[Y>0]==2)
  AUC <- calculate_AUC(curve_data$FPR, curve_data$TPR)
  results_null <- c(results_null, AUC)}, 
  error=function(cond) message(cond))
}
```

The p-value is calculated as proportion of differences with value equal to or less than zero.

```{r eval=FALSE}
summary(results_basic-results_null)
hist(results_basic-results_null, breaks=15, xlab = "Difference in AUROC", 
     main = "Basic vs. null models")
```
```{r echo=FALSE}
data("discrimination_model_validation")
summary(summary(discrimination_model_validation[[1]] - discrimination_model_validation[[2]]))
hist(discrimination_model_validation[[1]]-discrimination_model_validation[[2]], breaks=15, xlab = "Difference in AUROC",
     main = "Basic vs. null models")
```
All differences are positive, so the p-value is less than 0.001 since we have 1000 values.
