# Finding peaks differentiating callow and mature *F. sanguinea* workers

The procedure identifying species markers was also applied to determine peaks characteristic of *F. sanguinea* callow ants. In discriminant analysis, samples were classified into one of the three groups: callow *F. sanguinea*,  adult *F. sanguinea*, or adult *F. fusca*.

```{r warning=FALSE}
# devtools::install_github("mixOmicsTeam/mixOmics", ref = "143d31f9629bc83f1f27272ba6b7a2aaef0ee0e8")
suppressMessages({
  library(sangchem)
  library(dplyr)
  library(mixOmics)
})

# load data
data("development_data")
data("mass_spectra_data")

DA_data <- dplyr::filter(development_data, caste=="worker", !remarks %in% c("aggression actor", "aggression target", "queenless colony")) %>%
  dplyr::select(colony, species, callow, census_date, chromatogram_ID, sang_prop) %>%
  split(list(.$colony, .$census_date), drop=TRUE) %>%
  lapply(function(x){if(sum(x$callow)==0) return(NULL); x}) %>%
  {function(x) x[!unlist((lapply(x, is.null)))]}() %>%
  purrr::map2(seq_along(.), function(x,y){x$group <- y; x}) %>%
  {function(x) do.call(rbind, x)}()

Y <- (DA_data$callow==1) + (DA_data$species=="F. sanguinea")

peak_prop <- peak_proportions_table(mass_spectra_data[mass_spectra_data$chromatogram_ID %in% DA_data$chromatogram_ID,])
peak_prop <- peak_prop[match(DA_data$chromatogram_ID, rownames(peak_prop)),]
# remove absent peaks
peak_prop <- peak_prop[,apply(peak_prop,2,function(x) sum(x)>0)]
peak_prop <- freq_abun_QC(peak_prop, freq_cutoff = 0.3, abundance_cutoff = 0.005)
# replace zeros with a small number before transformation
peak_prop[peak_prop==0] <- 10^-5
# apply central log ratio transformation
peak_transformed <- t(apply(peak_prop, 1, clr_transformation))
PCA_callow_discr <- prcomp(peak_transformed, scale. = TRUE)
# PCA_callow_discr$x <- withinVariation(PCA_callow_discr$x, data.frame(DA_data$group))
n_PCs <- sum(cumsum(apply(PCA_callow_discr$x,2,var))/sum(apply(PCA_callow_discr$x,2,var))<0.8)+1
PCA_callow_discr$rotation <- PCA_callow_discr$rotation[,1:n_PCs]
PCA_callow_discr$x <- PCA_callow_discr$x[,1:n_PCs]
```
```{r}
set.seed(1342)
discr_analysis_callow <- splsda(PCA_callow_discr$x, Y=Y ,multilevel=DA_data$group, 
                                ncomp=7, scale = FALSE)
```

```{r fig.cap="Distribution of the samples projected onto two first latent components before optimizing the number of model paramaters."}
samples <- Y
samples[Y==0] = "mature F. fusca"
samples[Y==1] = "mature F. sanguinea"
samples[Y==2] = "callow F. sanguinea"
plotIndiv(discr_analysis_callow, legend=TRUE,ellipse=TRUE,comp=1:2, group=samples)
```

```{r}
perf.discr_analysis_callow <- perf(discr_analysis_callow, 
                                validation = "Mfold", 
                                folds = 4, 
                           progressBar = FALSE, 
                           auc = TRUE, 
                           nrepeat = 30,
                           scale = FALSE,
                           BPPARAM = BiocParallel::SerialParam(RNG=7320))
# plot(perf.discr_analysis_callow, col = color.mixo(1:3), sd = TRUE, legend.position = "horizontal")
selected_n <- perf.discr_analysis_callow$choice.ncom["overall", "max.dist"]
discr_analysis_callows_tuned <- tune.splsda(PCA_callow_discr$x,
                                  multilevel = DA_data$group,
                                   Y=Y,
                                   nrepeat = 30,
                                   folds = 4,
                                   ncomp = selected_n,
                                   test.keepX = seq(1:dim(PCA_callow_discr$x)[2]), 
                                   dist = 'max.dist',
                                   validation = 'Mfold',
                                   measure = "BER",
                                   progressBar = FALSE,
                                   scale = FALSE,
                                  BPPARAM = BiocParallel::SerialParam(RNG=7320)) 

selected_n <- discr_analysis_callows_tuned$choice.ncomp$ncomp
discr_analysis_callows_final <- splsda(PCA_callow_discr$x,Y=Y,
                      multilevel= DA_data$group,
                      keepX=discr_analysis_callows_tuned$choice.keepX[1:selected_n],
                      ncomp=selected_n,
                      scale = FALSE)
```

```{r fig.cap = "Projection of the samples in onto two first discriminant analysis components after model tuning."}
plotIndiv(discr_analysis_callows_final,
          legend=TRUE, ellipse=TRUE, comp=1:2, group = samples)
```

```{r}
rot_m <- PCA_callow_discr$rotation

Y.factor=discr_analysis_callows_final$Y
Y=discr_analysis_callows_final$ind.mat

X=list(X=discr_analysis_callows_final$X)
variatesX = discr_analysis_callows_final$variates$X
loadingsX = discr_analysis_callows_final$loadings$X

# Estimation matrix W, P and C
Pmat = crossprod(X[[1]], variatesX)
Cmat = crossprod(Y, variatesX)
Wmat = loadingsX

transf_matr = Wmat %*% solve(t(Pmat) %*% Wmat) %*% t(Cmat)

orig_var_effects <- t(t(transf_matr) %*% ginv(rot_m))
rownames(orig_var_effects) <- rownames(rot_m)

peaks_callow <- as.numeric(rownames(orig_var_effects)[orig_var_effects[,3]>=quantile(orig_var_effects[,3], 0.8)])
peaks_mature <- as.numeric(rownames(orig_var_effects)[orig_var_effects[,2]>=quantile(orig_var_effects[,2], 0.8)])
compounds <- c()
for(peak in rownames(orig_var_effects)){
  compounds <- c(compounds, mass_spectra_data[mass_spectra_data$peak_ID==peak,"identification"][1])
}
marker_scores_df <- data.frame(peak_ID = rownames(orig_var_effects), compound = compounds, callow_score = orig_var_effects[,3], callow_marker = rownames(orig_var_effects) %in% peaks_callow, mature_score=orig_var_effects[,2],
                               mature_marker = rownames(orig_var_effects) %in% peaks_mature)
marker_scores_df <- marker_scores_df[order(marker_scores_df$callow_score, decreasing = TRUE),]
rownames(marker_scores_df) <- NULL
```
```{r}
kableExtra::kbl(marker_scores_df, longtable=TRUE, booktabs = TRUE, caption = 'List of the peaks with the score indicating their importance as a marker of mature \\textit{F. fusca}, mature \\textit{F. sanguinea}, or callow \\textit{F. sanguinea} samples.',col.names = c("Peak ID", "Compound", "Callow score", "Callow marker", "Mature score", "Mature marker")) %>%
  kable_styling(latex_options = c("repeat_header"))
```
```{r eval=params$save_globals}
globals <- readRDS(system.file("globals.rds", package="sangchem"))
globals[["discr_analysis_callows_final"]] <- discr_analysis_callows_final
globals[["PCA_callow_discr"]] <- PCA_callow_discr
globals[["peaks_mature"]] <- as.numeric(peaks_mature)
globals[["peaks_callow"]] <- as.numeric(peaks_callow)
saveRDS(globals, system.file("globals.rds", package="sangchem"))
```

## Verifying the performace of the discrimination model

The performance of the discriminant model was evaluated by examining the accuracy of its predictions in a cross-validation test, measured as the area under the receiver operating characteristic curve (AUROC). This procedure was repeated $10^3$ times to account for variance resulting from the random splitting of samples into training and validation sets. In a parallel analysis, the model was trained on data with permuted age status of *F. sanguinea* to assess whether the true assignment of samples influenced model performance. Results from both training approaches were paired, and the difference in AUROC was calculated each for pair. The empirical p-value was defined as the proportion of models in which the AUROC after permutation was equal to or greater than the AUROC based on the original data.

```{r eval = FALSE}
# running basic model 
set.seed(1342)
results_basic <- c()
while(length(results_basic)<1000){
  message(counter)
  .Random.seed <- results$random.seed
  discr_analysis_callow <- splsda(PCA_callow_discr$x, Y=Y ,
                                  ncomp=7, scale = FALSE)
  tryCatch({perf.discr_analysis_callow <- perf(discr_analysis_callow,
                                               validation = "Mfold",
                                               folds = 4,
                                               progressBar = FALSE,
                                               auc = TRUE,
                                               nrepeat = 30,
                                               scale = FALSE,
                                               cpus=16)
  
  selected_n <- perf.discr_analysis_callow$choice.ncom["overall", "max.dist"]
  discr_analysis_callows_tuned <- tune.splsda(PCA_callow_discr$x,
                                              Y=Y,
                                              ncomp = selected_n,
                                              test.keepX = seq(1:120),
                                              validation = 'Mfold',
                                              folds = 4, 
                                              nrepeat = 30,
                                              dist = 'max.dist', # use max.dist measure
                                              measure = "BER",
                                              progressBar = FALSE,
                                              scale = FALSE)
  
  n_comp <- discr_analysis_callows_tuned$choice.ncomp$ncomp
  if(is.null(n_comp)) n_comp <- 1
  discr_analysis_callows_res <- splsda(PCA_callow_discr$x,Y=Y,
                                       keepX=discr_analysis_callows_tuned$choice.keepX,
                                       ncomp=n_comp,
                                       scale = FALSE)
  
  perf.discr_analysis_callows_res <- perf(discr_analysis_callows_res,
                                          validation = "Mfold",
                                          folds = 4,
                                          progressBar = FALSE,
                                          nrepeat = 1,
                                          scale = FALSE,
                                          cpus=16)
  
  curve_data <- calculate_accuracy_metrics(perf.discr_analysis_callows_res$predict[[length(perf.discr_analysis_callows_res$predict)]][,3,1][Y>0], Y[Y>0]==2)
  AUC <- calculate_AUC(curve_data$FPR, curve_data$TPR)
  results_basic <- c(results_basic, AUC)}, 
  error=function(cond) message(cond))
}


# running null models
set.seed(1342)
results_null <- c()
while(length(results_null)<1000){
  # shuffle categories 1 (mature F. sanguinea) and 2 (callow F. sanguinea)
  Y <- split(Y, DA_data$group) %>% lapply(function(x) {x[x>0]<-sample(x[x>0], length(x[x>0]));x}) %>% unlist()
  discr_analysis_callow <- splsda(PCA_callow_discr$x, Y=Y ,
                                  ncomp=7, scale = FALSE)
  tryCatch({perf.discr_analysis_callow <- perf(discr_analysis_callow,
                                               validation = "Mfold",
                                               folds = 4,
                                               progressBar = FALSE,
                                               auc = TRUE,
                                               nrepeat = 30,
                                               scale = FALSE,
                                               cpus=16)
  
  selected_n <- perf.discr_analysis_callow$choice.ncom["overall", "max.dist"]
  discr_analysis_callows_tuned <- tune.splsda(PCA_callow_discr$x,
                                              Y=Y,
                                              ncomp = selected_n,
                                              test.keepX = seq(1:120),
                                              validation = 'Mfold',
                                              folds = 4, 
                                              nrepeat = 30,
                                              dist = 'max.dist', # use max.dist measure
                                              measure = "BER",
                                              progressBar = FALSE,
                                              scale = FALSE,
                                              cpus=16)
  
  n_comp <- discr_analysis_callows_tuned$choice.ncomp$ncomp
  if(is.null(n_comp)) n_comp <- 1
  discr_analysis_callows_res <- splsda(PCA_callow_discr$x,Y=Y,
                                       keepX=discr_analysis_callows_tuned$choice.keepX,
                                       ncomp=n_comp,
                                       scale = FALSE)
  
  perf.discr_analysis_callows_res <- perf(discr_analysis_callows_res,
                                          validation = "Mfold",
                                          folds = 4,
                                          progressBar = FALSE,
                                          nrepeat = 1,
                                          scale = FALSE,
                                          cpus=16)
  
  curve_data <- calculate_accuracy_metrics(perf.discr_analysis_callows_res$predict[[length(perf.discr_analysis_callows_res$predict)]][,3,1][Y>0], Y[Y>0]==2)
  AUC <- calculate_AUC(curve_data$FPR, curve_data$TPR)
  results_null <- c(results_null, AUC)}, 
  error=function(cond) message(cond))
}
```

The p-value is calculated as proportion of differences with value equal to or less than zero.

```{r eval=FALSE}
summary(results_basic-results_null)
hist(results_basic-results_null, breaks=15, xlab = "Difference in AUROC", 
     main = "Basic vs. null models")
```


```{r echo=FALSE, fig.cap="Distribution of the differences in the performance of the discriminant models trained on true and permuted data. Age status of \\textit{F. sanguinea} ants were shuffled in the null variant."}
data("discrimination_model_validation")
# summary(discrimination_model_validation[[1]] - discrimination_model_validation[[2]])
hist(discrimination_model_validation[[1]]-discrimination_model_validation[[2]], breaks=15, xlab = "Difference in AUROC",
     main = "Basic vs. null models")
```

All differences are positive, so the *p*-value is less than 0.001 since the null distribution consists of 1000 values.
