# Difference in the total mass between the species
```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(dplyr)
  library(lme4)
  library(lmerTest)
  library(rstatix)
  library(ggpubr)
})

# load data
data("development_data")
```
## Difference in the total CHC amount between mature *F. sanguinea* and *F. fusca* 

### CHC mass difference between heterospecific nestmates
```{r}
development_samples <- filter(development_data, caste=="worker", !remarks %in% c("aggression actor", "aggression target", "queenless colony")) %>% 
  mutate(corrected_mass=mass/head_width^2*10^6) %>%
  select(species, census_date, colony, corrected_mass,callow, chromatogram_ID)
test_data <- filter(development_samples,callow == 0) 
test_data_sang <- filter(test_data, species == "F. sanguinea") %>% colony_weighted_mean()
test_data_sang$species <- "F. sanguinea"
test_data_fusca <- filter(test_data, species == "F. fusca") %>% colony_weighted_mean()
test_data_fusca$species <- "F. fusca"
colonies <- intersect(test_data_sang$colony, test_data_fusca$colony)
test_df <- rbind(test_data_sang, test_data_fusca)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]

test_df$species <- factor(test_df$species)
test_stats <- wilcox_test(test_df, corrected_mass~species, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="species", dodge=0.8, fun="max")
ggboxplot(test_df, x="species", y="corrected_mass")+
  ylab("Corrected CHC mass in F. sanguinea and their F. fusca slaves")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)

```

### Difference in CHC mass predicted by linear model

The bootMer function enebales sampling predicted values based on the linear model. Here we will set F. sagnuinea proportion at 0 or 1 to predict CHC mass for pure F. fusca and F. sanguinea colonies, respecitvely. Then we shall calculate sapling distribution of the differences between 
```{r eval = FALSE}
# load data
dev_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==0) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(corrected_mass=mass/head_width^2*10^6)

model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop  + (1|colony) +(1 | colony:census_date), data=model_input)

simulation_number=10^6
set.seed(4512)
sang_fam_simulated<-bootMer(lm_model,function(x) predict(x,data.frame(sang_prop=1),re.form=NA),nsim=simulation_number)$t
pred_values_sang <- exp(sang_fam_simulated)


model_input <- dev_samples[dev_samples$species=="F. fusca",]
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop  + (1|colony), data=model_input)
fus_fam_simulated<-bootMer(lm_model,function(x) predict(x,data.frame(sang_prop=0),re.form=NA),nsim=simulation_number)$t
pred_values_fusca <- exp(fus_fam_simulated)
```
```{r echo=FALSE}
data("predicted_CHC_mass")
pred_values_sang <- predicted_CHC_mass[[1]]
pred_values_fusca <- predicted_CHC_mass[[2]]
```
```{r}
summary(pred_values_sang-pred_values_fusca)
```

