# Changes of CHCs amount over time

This chapter contains a series of linear models fitted to examine how the proportion of *F. sanguinea* workers in a colony is associated with the amount of CHC, analyzed together or in subsets.

```{r}
suppressMessages({
library(sangchem)
library(dplyr)
library(lme4)
library(lmerTest)
library(DHARMa)
})
# load data
data("development_data")
load_globals()
dev_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==0) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(corrected_mass=mass/head_width^2*10^6)
```

## Change in the total CHCs mass
### Mature *F. sanguinea*
```{r}
set.seed(3453)
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop  + (1|colony) +(1 | colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### *F. fusca*
```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop  + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### Callow *F. sanguinea*
```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(corrected_mass=mass/head_width^2*10^6)

lm_model <- lmer(I((corrected_mass)) ~ sang_prop  + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

## Change of the CHCs associated with *F. sanguinea*
### Mature *F. sanguinea* ants
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lmer(I(log(corrected_mass+0.1)) ~ sang_prop +(1 | colony:census_date) + (1|colony), data=model_input)
print(summary(lm_model))
# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### *F. fusca* ants
```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lmer(I(sqrt(corrected_mass)) ~ sang_prop +(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### Callow *F. sanguinea* ants
```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(corrected_mass=mass/head_width^2*10^6)
# random term dropped because of the model singularity
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lm(I((corrected_mass)^(1/3)) ~ sang_prop, data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
```

## Change of the CHCs associated with *F. fusca*
### Mature *F. sanguinea* ants
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(corrected_mass+1)) ~ sang_prop +(1 | colony:census_date) + (1 | colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n=1e4)
plot(simulationOutput)
```

### *F. fusca* ants
```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
model_input$corrected_mass <- model_input$corrected_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(corrected_mass)) ~ sang_prop + (1 | colony)+ (1 | colony:census_date), data=model_input)
print(summary(lm_model))

simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n=1e4)
plot(simulationOutput)

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n=1e4)
plot(simulationOutput)
```
