# Data exploration

This chapter presents general overview of the data with the calculations whose results are presented in the manuscript as summary statistics.


## General results
```{r}
suppressMessages({
  library(sangchem)
  library(dplyr)
  library(stringi)
})

# load data from the the earlier study
data("field_colonies")
data("development_data")
data("mass_spectra_data")

cat(sprintf("Number of peaks: %d\n", length(setdiff(mass_spectra_data$peak_ID, 0:1))))
```


## Proportion of the marker CHCs in the total CHC signal
```{r}
# load marker peak IDs
load_globals()

cat(sprintf("Number of fusca markers: %d\n", length(peaks_fusca)))
cat(sprintf("Number of sanguinea markers: %d\n", length(peaks_sang)))
# select samples from F. sanguinea ants and free-living F. fusca ants encoded in chromatogram_ID
fusca_samples <- as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID), "fus[0-9].?-[0-9]\\([1-2]\\)")])
sanguinea_samples<-as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID),"sangFs[0-9].?\\([0-9]\\)")])

# remove outliers
sanguinea_samples <- setdiff(sanguinea_samples,c("sangFs2(1)","sangFs5(2)","sangFs26(1)","sangFs26(2)"))
fusca_samples <- setdiff(fusca_samples,c("fus9-2(1)", "fus26-3(2)"))

sang_markers_in_fusca <- c()
fus_markers_in_fusca <- c()
for(fus_sample in fusca_samples){
  sample_data <- field_colonies[field_colonies$chromatogram_ID==fus_sample,]
  total_signal <- sum(sample_data[sample_data$peak_ID > 1, "abundance"])
  sang_markers_in_fusca <- c(sang_markers_in_fusca, sum(sample_data[sample_data$peak_ID %in% peaks_sang, "abundance"])/total_signal)
  fus_markers_in_fusca <- c(fus_markers_in_fusca, sum(sample_data[sample_data$peak_ID %in% peaks_fusca, "abundance"])/total_signal)
}
cat(sprintf("Mean proportion of fusca markers in fusca workers: %.3f\n", round(mean(fus_markers_in_fusca), digits = 3)))
cat(sprintf("Mean proportion of sanguinea markers in fusca workers: %.3f\n", round(mean(sang_markers_in_fusca), digits = 3)))

sang_markers_in_sanguinea <- c()
fus_markers_in_sanguinea <- c()
for(sang_sample in sanguinea_samples){
  sample_data <- field_colonies[field_colonies$chromatogram_ID==sang_sample,]
  total_signal <- sum(sample_data[sample_data$peak_ID > 1, "abundance"])
  sang_markers_in_sanguinea <- c(sang_markers_in_sanguinea, sum(sample_data[sample_data$peak_ID %in% peaks_sang, "abundance"])/total_signal)
  fus_markers_in_sanguinea <- c(fus_markers_in_sanguinea, sum(sample_data[sample_data$peak_ID %in% peaks_fusca, "abundance"])/total_signal)
}
cat(sprintf("Mean proportion of fusca markers in sanguinea workers: %.3f\n", round(mean(fus_markers_in_sanguinea), digits = 3)))
cat(sprintf("Mean proportion of sanguinea markers in sanguinea workers: %.3f\n", round(mean(sang_markers_in_sanguinea), digits = 3)))

```

## CHCs mean absolute amount

The mass is standarized so that it is divided by square of the ratio between true head with and 1.3 mm.

```{r}

data("development_data")
data("body_surface_area")
CHC_normalizer <- get_body_area_normalizer(body_surface_area, development_data)
sang_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), species=="F. sanguinea") %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date,callow) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width))

cat(sprintf("Mean CHC mass for callow workers: %f, standard deviation: %f\n", mean(filter(sang_samples, callow==1)$normalized_mass), sd(filter(sang_samples, callow==1)$normalized_mass)))
cat(sprintf("Mean CHC mass for mature workers: %f, standard deviation: %f\n", mean(filter(sang_samples, callow==0)$normalized_mass), sd(filter(sang_samples, callow==0)$normalized_mass)))
```


## Numbers of samples
```{r}
dev_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony")) %>%
  dplyr::select(species, colony, census_date,callow, sang_prop) 
cat(sprintf("The number of samples representing F. sanguinea mature ants: %d\n", nrow(filter(dev_samples, species == "F. sanguinea", callow == 0))))
cat(sprintf("The number of samples representing F. sanguinea callow ants: %d\n", nrow(filter(dev_samples, species == "F. sanguinea", callow == 1))))
cat(sprintf("The number of samples representing F. fusca ants: %d\n", nrow(filter(dev_samples, species == "F. fusca"))))
cat(sprintf("The range of F. sanguinea proportions: %s\n", paste(as.character(range(dev_samples$sang_prop)), collapse = "-")))
cat(sprintf("Number of samples from colonies with no slave-maker workers: %d\n", nrow(filter(dev_samples, sang_prop == 0))))
cat(sprintf("Number of samples from colonies with no slaves: %d\n", nrow(filter(dev_samples, sang_prop == 1))))
```


## Proportion of docosane
```{r}
data("separation_experiment_data")
data("mass_spectra_data")
control_treatment <- separation_experiment_data[separation_experiment_data$treatment_id==9, "chromatogram_ID"]
cat("Proportion of C22 of intrinsic origin")
summary(unlist(lapply(control_treatment, function(x) peak_proportion(1, x, mass_spectra_data))))
control_treatment <- separation_experiment_data[separation_experiment_data$treatment_id==5, "chromatogram_ID"]
cat("Proportion of C22 in ants affected by the glass beads with F. fusca CHCs")
summary(unlist(lapply(control_treatment, function(x) peak_proportion(1, x, mass_spectra_data))))
control_treatment <- separation_experiment_data[separation_experiment_data$treatment_id==7, "chromatogram_ID"]
cat("Proportion of C22 in ants affected by the glass beads with F. sanguinea CHCs")
summary(unlist(lapply(control_treatment, function(x) peak_proportion(1, x, mass_spectra_data))))
```


## Session info
```{r}
print(sessionInfo())
```
