# Effect of dummy ants

```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(dplyr)
})

# load data
data("separation_experiment_data")
data("mass_spectra_data")
data("subject_to_donor_mapping")
```

## Distance to the CHC profile of the dummy ants treated by F. fusca CHCs

We computed the chemical distance to the CHC profile of ants whose CHCs were used to cover glass beads serving as dummy ants.
```{r}
test_dummy_ant_effect(treatment_id=5, separation_experiment_data, mass_spectra_data, subject_to_donor_mapping)
```

## Distance to the CHC profile of the dummy ants treated by F.sanguinea CHCs
```{r}
test_dummy_ant_effect(treatment_id=7, separation_experiment_data, mass_spectra_data, subject_to_donor_mapping)
```

## Fraction of docosane in the CHC profile

For control treatment, glass beads were not covered by CHCs. In experimental treatment CHCs from donor colonies were contaminated with docosane which served as internal standard. We look for the differences in the proportion on docosane in the CHC profile of tested ants to judge about chemicals acquisition from dummy ants.

### F. fusca

```{r}
test_C22_proportion(5, separation_experiment_data, mass_spectra_data)
```

### F. sanguinea

```{r}
test_C22_proportion(7, separation_experiment_data, mass_spectra_data)
```
