# Comparison of the amount and proportions of CHC between species and age categories

This section presents the results of Wilcoxon tests comparing features of CHC profiles between callow and mature *F. sanguinea* ants as well as *F. fusca* slaves. Samples from the same colony were averaged before calculation of the final statistics to account for their non-independence.  

```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(dplyr)
  library(ggpubr)
  library(rstatix)
})

# load data
data("development_data")
data("mass_spectra_data")
development_samples <- filter(development_data, caste=="worker", !remarks %in% c("aggression actor", "aggression target", "queenless colony")) %>% 
  mutate(corrected_mass=mass/head_width^2*10^6) %>%
  select(species, census_date, colony, corrected_mass,callow, chromatogram_ID)
load_globals()
```

## Difference in the total CHC amount between *F. sanguinea* and *F. fusca*
```{r}
test_data <- filter(development_samples, callow == 0) 
test_data_sang <- filter(test_data,  species=="F. sanguinea") %>% colony_weighted_mean()
test_data_sang$species <- "F. sanginea"
test_data_fusca <- filter(test_data, species == "F. fusca") %>% colony_weighted_mean()
test_data_fusca$species <- "F. fusca"
colonies <- intersect(test_data_sang$colony, test_data_fusca$colony)
test_df <- rbind(test_data_sang, test_data_fusca)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
wilcox_test(test_df, corrected_mass~species, paired=TRUE)
test_df$species <- factor(test_df$species)
test_stats <- wilcox_test(test_df, corrected_mass~species, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="species", dodge=0.8, fun="max")
ggboxplot(test_df, x="species", y="corrected_mass")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```

## Difference in the total CHC amount between callow and mature *F. sanguinea*
```{r}
test_data <- filter(development_samples, species == "F. sanguinea") 
test_data_mature <- filter(test_data, callow == 0) %>% colony_weighted_mean()
test_data_mature$callow <- 0
test_data_callow <- filter(test_data, callow == 1) %>% colony_weighted_mean()
test_data_callow $callow <- 1
colonies <- intersect(test_data_mature$colony, test_data_callow$colony)
test_df <- rbind(test_data_callow, test_data_mature)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
wilcox_test(test_df, corrected_mass~callow, paired=TRUE)
test_df$callow <- factor(test_df$callow)
test_stats <- wilcox_test(test_df, corrected_mass~callow, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="callow", dodge=0.8, fun="max")
ggboxplot(test_df, x="callow", y="corrected_mass")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```

## Difference in the proportion of CHC characteristic of *F. sanguinea* between mature *F. sanguinea* and *F. fusca*
```{r}
test_data <- development_samples
test_data$sang_prop <- subsample_proportion(peaks_sang, test_data$chromatogram_ID)
test_data_sang <- filter(test_data, species == "F. sanguinea", callow ==  0) %>% colony_weighted_mean(var = "sang_prop")
test_data_sang$species <- "F. sanguinea"
test_data_fusca <- filter(test_data, species == "F. fusca") %>% colony_weighted_mean(var = "sang_prop")
test_data_fusca$species <- "F. fusca"
colonies <- intersect(test_data_sang$colony, test_data_fusca$colony)
test_df <- rbind(test_data_fusca, test_data_sang)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$species <- factor(test_df$species)
test_stats <- wilcox_test(test_df, sang_prop~species, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="species", dodge=0.8, fun="max")
ggboxplot(test_df, x="species", y="sang_prop")+
  ylab("Proportion of CHC characteristic of F. sanguinea")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```

## Difference in the proportion of CHC characteristic of *F. sanguinea* between mature and callow *F. sanguinea* 
```{r}
test_data <- filter(development_samples, species == "F. sanguinea")
test_data$sang_prop <- subsample_proportion(peaks_sang, test_data$chromatogram_ID)
test_data_mature <- filter(test_data, callow ==  0) %>% 
  colony_weighted_mean(var = "sang_prop")
test_data_mature$callow <- 0
test_data_callow <- filter(test_data,callow == 1) %>% colony_weighted_mean(var = "sang_prop")
test_data_callow$callow <- 1
colonies <- intersect(test_data_mature$colony, test_data_callow$colony)
test_df <- rbind(test_data_mature, test_data_callow)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$callow <- factor(test_df$callow)
test_stats <- wilcox_test(test_df, sang_prop~callow, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="callow", dodge=0.8, fun="max")
ggboxplot(test_df, x="callow", y="sang_prop")+
  ylab("Proportion of CHC characteristic of F. sanguinea between mature and callow F. sanguinea")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```

## Difference in the proportion of CHC characteristic of callow *F. sanguinea* between mature and callow *F. sanguinea* 
```{r}
test_data <- filter(development_samples, species == "F. sanguinea")
test_data$sang_prop <- subsample_proportion(peaks_callow, test_data$chromatogram_ID)
test_data_mature <- filter(test_data, callow ==  0) %>% 
  colony_weighted_mean(var = "sang_prop")
test_data_mature$callow <- 0
test_data_callow <- filter(test_data,callow == 1) %>% colony_weighted_mean(var = "sang_prop")
test_data_callow$callow <- 1
colonies <- intersect(test_data_mature$colony, test_data_callow$colony)
test_df <- rbind(test_data_mature, test_data_callow)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$callow <- factor(test_df$callow)
test_stats <- wilcox_test(test_df, sang_prop~callow, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="callow", dodge=0.8, fun="max")
ggboxplot(test_df, x="callow", y="sang_prop")+
  ylab("Proportion of CHC characteristic of callow F. sanguinea\nbetween mature and callow F. sanguinea")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```


## Difference in the amount of CHC characteristic of callow *F. sanguinea* between mature and callow *F. sanguinea* 
```{r}
test_data <- filter(development_samples, species == "F. sanguinea")
test_data$marker_prop <- subsample_proportion(peaks_callow, test_data$chromatogram_ID)
test_data$marker_mass <- test_data$marker_prop * test_data$corrected_mass
test_data_mature <- filter(test_data, callow ==  0) %>% 
  colony_weighted_mean(var = "marker_mass")
test_data_mature$callow <- 0
test_data_callow<- filter(test_data, callow == 1) %>% colony_weighted_mean(var = "marker_mass")
test_data_callow$callow <- 1
colonies <- intersect(test_data_mature$colony, test_data_callow$colony)
test_df <- rbind(test_data_mature, test_data_callow)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$callow <- factor(test_df$callow)
test_stats <- wilcox_test(test_df, marker_mass~callow, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="callow", dodge=0.8, fun="max")
ggboxplot(test_df, x="callow", y="marker_mass")+
  ylab(" Difference in the mass of callow markers between callow and mature F. sanguinea")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```

## Difference in the proportion of CHC characteristic of *F. fusca* between mature *F. sanguinea* and *F. fusca*
```{r}
test_data <- development_samples
test_data$fusca_prop <- subsample_proportion(peaks_fusca, test_data$chromatogram_ID)
test_data_sang <- filter(test_data, species == "F. sanguinea", callow ==  0) %>% colony_weighted_mean(var = "fusca_prop")
test_data_sang$species <- "F. sanguinea"
test_data_fusca <- filter(test_data, species == "F. fusca") %>% colony_weighted_mean(var = "fusca_prop")
test_data_fusca$species <- "F. fusca"
colonies <- intersect(test_data_sang$colony, test_data_fusca$colony)
test_df <- rbind(test_data_fusca, test_data_sang)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$species <- factor(test_df$species)
test_stats <- wilcox_test(test_df, fusca_prop~species, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="species", dodge=0.8, fun="max")
ggboxplot(test_df, x="species", y="fusca_prop")+
  ylab("Proportion of CHC characteristic of F. fusca")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```

## Difference in the proportion of CHC characteristic of *F. fusca* between mature and callow *F. sanguinea* 
```{r}
test_data <- filter(development_samples, species == "F. sanguinea")
test_data$fusca_prop <- subsample_proportion(peaks_fusca, test_data$chromatogram_ID)
test_data_mature <- filter(test_data, callow ==  0) %>% 
  colony_weighted_mean(var = "fusca_prop")
test_data_mature$callow <- 0
test_data_callow <- filter(test_data,callow == 1) %>% colony_weighted_mean(var = "fusca_prop")
test_data_callow$callow <- 1
colonies <- intersect(test_data_mature$colony, test_data_callow$colony)
test_df <- rbind(test_data_mature, test_data_callow)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$callow <- factor(test_df$callow)
test_stats <- wilcox_test(test_df, fusca_prop~callow, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="callow", dodge=0.8, fun="max")
ggboxplot(test_df, x="callow", y="fusca_prop")+
  ylab("Proportion of CHC characteristic of F. fusca in mature and callow F. sanguinea")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```


## Difference in the proportion of CHC characteristic of *F. fusca* between callow *F. sanguinea* and *F. fusca*
```{r}
test_data <- filter(development_samples)
test_data$fusca_prop <- subsample_proportion(peaks_fusca, test_data$chromatogram_ID)
test_data_sang <- filter(test_data, callow ==  1, species =="F. sanguinea") %>% 
  colony_weighted_mean(var = "fusca_prop")
test_data_sang$species <- "F. sanguinea"
test_data_fusca <- filter(test_data,species == "F. fusca") %>% colony_weighted_mean(var = "fusca_prop")
test_data_fusca$species <- "F. fusca"
colonies <- intersect(test_data_fusca$colony, test_data_sang$colony)
test_df <- rbind(test_data_sang, test_data_fusca)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$species <- factor(test_df$species)
test_stats <- wilcox_test(test_df, fusca_prop~species, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="species", dodge=0.8, fun="max")
ggboxplot(test_df, x="species", y="fusca_prop")+
  ylab("Proportion of CHC characteristic of F. fusca in F. fusca callow F. sanguinea")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```

## Difference in the mass of *n*-alkanes between callow and mature *F. sanguinea*
```{r}
n_alkanes_ID <- unique(mass_spectra_data$peak_ID[grep("^C[0-9]{2}$",mass_spectra_data$identification)])
test_data <- filter(development_samples, species == "F. sanguinea")
test_data$n_alkanes_prop <- subsample_proportion(n_alkanes_ID, test_data$chromatogram_ID)
test_data$n_alkanes_mass <- test_data$n_alkanes_prop * test_data$corrected_mass
test_data_mature <- filter(test_data, callow ==  0) %>% 
  colony_weighted_mean(var = "n_alkanes_mass")
test_data_mature$callow <- 0
test_data_callow<- filter(test_data, callow == 1) %>% colony_weighted_mean(var = "n_alkanes_mass")
test_data_callow$callow <- 1
colonies <- intersect(test_data_mature$colony, test_data_callow$colony)
test_df <- rbind(test_data_mature, test_data_callow)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$callow <- factor(test_df$callow)
test_stats <- wilcox_test(test_df, n_alkanes_mass~callow, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="callow", dodge=0.8, fun="max")
ggboxplot(test_df, x="callow", y="n_alkanes_mass")+
  ylab(" Difference in the mass of n-alkanes between callow and mature F. sanguinea")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```

## Difference in the proprtion of *n*-alkanes between callow and mature *F. sanguinea*
```{r}
test_data <- filter(development_samples, species == "F. sanguinea")
test_data$n_alkanes_prop <- subsample_proportion(n_alkanes_ID, test_data$chromatogram_ID)
test_data_mature <- filter(test_data, callow ==  0) %>% 
  colony_weighted_mean(var = "n_alkanes_prop")
test_data_mature$callow <- 0
test_data_callow<- filter(test_data, callow == 1) %>% colony_weighted_mean(var = "n_alkanes_prop")
test_data_callow$callow <- 1
colonies <- intersect(test_data_mature$colony, test_data_callow$colony)
test_df <- rbind(test_data_mature, test_data_callow)
test_df <- test_df[test_df$colony %in% colonies,]
test_df <- test_df[order(test_df$colony),]
test_df$callow <- factor(test_df$callow)
test_stats <- wilcox_test(test_df, n_alkanes_prop~callow, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="callow", dodge=0.8, fun="max")
ggboxplot(test_df, x="callow", y="n_alkanes_prop")+
  ylab(" Difference in the proportion of n-alkanes between callow and mature F. sanguinea")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(test_stats, label="p", tip.length = 0, size=3.5)
```
