# Species discrimination and prediction

## Species markers
This  report shows how the peaks discriminative for *F. sanguinea* and *F. fusca* were determined. We will use the chemical data collected from the filed colonies in the earlier study [@WLODARCZYK201798]. In this way our method of markers selection is blind to the data set to which  markers will be applied avoiding potential source of the bias in the analysis results.
```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(stringi)
  library(dplyr)
  library(mixOmics)
})

# load data from the the earlier study
data("field_colonies")

# select samples from F. sanguinea ants and free-living F. fusca ants encoded in chromatogram_ID
fusca_samples <- as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID), "fus[0-9].?-[0-9]\\([1-2]\\)")])
sanguinea_samples<-as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID),"sangFs[0-9].?\\([0-9]\\)")])

# remove outliers
sanguinea_samples <- setdiff(sanguinea_samples,c("sangFs2(1)","sangFs5(2)","sangFs26(1)","sangFs26(2)"))
fusca_samples <- setdiff(fusca_samples,c("fus9-2(1)", "fus26-3(2)"))
discrimination_data <- filter(field_colonies,chromatogram_ID %in% c(fusca_samples,sanguinea_samples)) 
discrimination_data <- filter(discrimination_data, peak_ID<110)
peak_prop_unfiltered <- peak_proportions_table(discrimination_data)
# feature selection
peak_prop <- freq_abun_QC(peak_prop_unfiltered)
# re-normalize rows to 1
peak_prop <- peak_prop/rowSums(peak_prop)
# remove absent peaks
peak_prop <- peak_prop[,apply(peak_prop,2,function(x) sum(x)>0)]
# replace zeros with a small number before transformation
peak_prop[peak_prop==0] <- 10^-16
# apply central log ratio transformation
peak_transformed <- t(apply(peak_prop, 1, clr_transformation))
```

```{r}
PCA_species_discr <- prcomp(peak_transformed, scale. = TRUE)
```

We also need to define our response variable to be explained by discrimination model.
```{r}
#define groups to be compared
Y <- rownames(PCA_species_discr$x) %in% sanguinea_samples
```

Now that we have prepared the input data we can run the model. The following workflow is based on the tutorial available at the official [`mixOmics` package web site](https://mixomics.org/mixmc/hmp-bodysites-case-study/).
```{r}
set.seed(9230)
# run the main model
species_discrimination_PC.plsda <- plsda(PCA_species_discr$x, Y, ncomp=10, scale=FALSE)
plotIndiv(species_discrimination_PC.plsda,group=Y,
          legend=TRUE,ellipse=TRUE,comp=1:2)

# determine the optimal number of components used in the model
spec_discr_PC.plsda.perf<-perf(species_discrimination_PC.plsda, nrepeat = 10,folds=5,
                            progressBar = FALSE,validation = "Mfold",
                            auc=TRUE, scale=FALSE)

plot(spec_discr_PC.plsda.perf, sd = TRUE, overlay = 'measure')
```

We use the performance test result to determine the optimal number of components to be used in 
discriminant analysis.
```{r}
selected_n <- spec_discr_PC.plsda.perf$choice.ncom["overall", "max.dist"]

# determine the number of the original variables to be kept in each component (actually
# here we have only one component)
spec_discr_PC.splsda <- tune.splsda(PCA_species_discr$x,Y,test.keepX = c(1:dim(PCA_species_discr$x)[2]),ncomp = selected_n, validation="Mfold", progressBar=FALSE, dist="max.dist",
                                    nrepeat=5, cpus=2, folds=5, measure="BER", scale=FALSE)

# final model
species_prediction_model <- splsda(PCA_species_discr$x, Y, ncomp=selected_n,
                           keepX =spec_discr_PC.splsda$choice.keepX, scale=FALSE)

# visualize results for the first two components
plotIndiv(species_prediction_model, group=Y, legend=TRUE, ellipse=TRUE, comp=1:2)

```

Now we want assess the importance of each peak in the discrimination between the species. For that we generate 10^5 mock profiles which are transformed in the same manner as the training data. For each profile we calculate the prediction score. Values closer to one or greater indicate the CHC profile being characteristic of *F. sanguinea*. Then, for each peak, we use its value after transformation as a regressor of inferred species (prediction score). This approach is justified by the linearity of the relation between both variables. 
```{r message=FALSE}
set.seed(2341)
n_obs <- 10^5
n_var <- dim(peak_prop)[2]
x <- exp(5*runif(n_obs * n_var))
#x <- runif(n_obs * n_var)
m = matrix(x, nrow = n_obs, ncol=n_var)
m <- m/rowSums(m)
transformed_m <- t(apply(m, 1, clr_transformation))
# log_m <- apply(m,2,log)
colnames(m)  <- colnames(peak_prop)
colnames(transformed_m) <- paste0("peak_", colnames(m))
df <- as.data.frame(transformed_m)
df$y <- calculate_SII(m,  species_prediction_model, PCA_species_discr, predicted_category=2)[,2]
```
```{r}
set.seed(4563)
p_val = c()
peak_coefficients <- c()
selected_residuals <- round(seq(1,n_obs,length.out=3000))
for(peak in colnames(transformed_m)){
  print(peak)
  model <- lm(formula(paste0("y~1 + ", peak)), data =df)
  print(summary(model))
  hist(residuals(model), breaks = 50, main = paste0("residuals ", peak))
  print(shapiro.test(sort(residuals(model))[selected_residuals]))
  peak_coefficients <- c(peak_coefficients, model$coefficients[2])
  plot(model)

}
```
```{r}
compounds <- c()
for (peak in colnames(peak_prop)){
  compounds <- c(compounds, mass_spectra_data[mass_spectra_data$peak_ID==peak,"identification"][1])
}
df_importance <- data.frame(peak_ID=colnames(peak_prop), coefficient = peak_coefficients, identification=compounds) 
print(df_importance[order(df_importance$coefficient),])
```


```{r}
peak_fusca <- as.numeric(gsub("peak_", "", names(peak_coefficients)[peak_coefficients<(-10^-3)]))
peak_sang <- as.numeric(gsub("peak_", "", names(peak_coefficients)[peak_coefficients>(10^-3)]))
``` 
The objects `peaks_sang`, `peaks_fusca`, `species_prediction_model` and `PCA_species_discr` are stored in the package and can be collectively loaded into the global environment along with the other objects by running `load_globals()`.
```{r eval=FALSE}
globals <- readRDS(system.file("globals.rds", package="sangchem"))
globals[["species_prediction_model"]] <- species_prediction_model
globals[["PCA_species_discr"]] <- PCA_species_discr
globals[["peak_sang"]] <- peaks_sang
globals[["peak_fusca"]] <- peaks_fusca
saveRDS(globals, system.file("globals.rds", package="sangchem"))
```

## Species identity score
Ants in mixed colonies mixed CHCs to produce hybrid chemical identity. Here we develop the method to assign each sample a number indicating how much it is characteristic of one or the other species.
The main idea is to fit a sPLS_DA model to the training set (sampled from field colonies) and use it predict the explanatory variable (species) in the test data (samples from experimental colonies). However the sPLS_DA model run on the raw data will yield the classificator which used only few variables (peaks) due to the well separation of samples of both species. The unused variables might also be informative in the species discrimination task. We want to use as much data as it is possible to make robust predictions and thus we will use principal component analysis output. Thus, the rotation matrix from the principal component analysis will be used to transform input data to the discrimination model. The correspondig code in encapsulated in the `claculate_SII` function.

### Data preparation
```{r}
data("development_data")
dev_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony")) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, callow, census_date)
normalized_abundances <- peak_proportions_table(mass_spectra_data[mass_spectra_data$chromatogram_ID %in% dev_samples$chromatogram_ID,])
species_indices <- calculate_SII(normalized_abundances)
dev_samples <- left_join(dev_samples, species_indices)
```

### Predicted species of different categories of ants as a function of the proportion of *F. sanguinea* ants in a colony
#### Mature *F. sanguinea*
```{r}
library(lmerTest)
library(DHARMa)
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==0,]
lm_model <- lmer(predicted_species~sang_prop+(1|colony)+(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

#### *F. fusca*
```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
lm_model <- lmer(predicted_species~sang_prop+(1|colony)+(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

#### Callow *F. sanguinea*
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==1,]
lm_model <- lmer(predicted_species~sang_prop+(1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

### Difference in species prediction between mature *F. sanguinea* ants and their slaves
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==0,]
model_input$index_diff <- model_input$predicted_species - find_nestmates_SII(model_input$chromatogram_ID, dev_samples)
lm_model <- lmer(index_diff~sang_prop+(1|colony)+(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE)
plot(simulationOutput)
```

### Difference in species prediction between callow *F. sanguinea* ants and their slaves

Since the linear model does not meet one of diagnostics criterion we will apply Wilcoxon paired test.
```{r message=FALSE}
library(ggpubr)
library(rstatix)
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==1,]
model_input$slave_species_index <- find_nestmates_SII(model_input$chromatogram_ID, dev_samples)
print(wilcox.test(model_input$predicted_species, model_input$slave_species_index, paired=TRUE))
df <- data.frame(species_identity_index=c(model_input$predicted_species, model_input$slave_species_index), species="F. fusca")
df$species[1:nrow(model_input)] <- "callow F. sanguinea"
stat_tests <- wilcox_test(df, species_identity_index~species, paired=TRUE) %>%
  add_significance("p") %>%
  add_xy_position(x="species", dodge=0.8, fun="max")
ggboxplot(df, x="species", y="species_identity_index")+
  scale_y_continuous(expand = expansion(mult=c(0.02,0.1)))+
  stat_pvalue_manual(stat_tests, label="p", tip.length = 0, size=3.5)
```

### Difference in species prediction between callow and mature *F. sanguinea* ants.
```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea"&dev_samples$callow==1,]
model_input$index_diff <- model_input$predicted_species - find_nestmates_SII(model_input$chromatogram_ID, dev_samples, heterospecific = FALSE)
# no random effect since it captures no variance
lm_model <- lm(index_diff~sang_prop, data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model))
print(shapiro.test(residuals(lm_model)))
```

## Session info
```{r}
print(sessionInfo())
```
