---
title: "Coming out from the shadows: facultative slave-making ants reveal their chemical identity during colony development"
subtitle: "Behavioral Ecology and Sociobiology"
author:
  - Tomasz Włodarczyk^[Faculty of Biology, University of Białystok, Poland, tomwlo@gmail.com, ORCID 0000-0003-1554-9699]
  - Thomas Schmitt^[Department of Animal Ecology and Tropical Biology, University of Würzburg, Germany, thomas.schmitt@uni-wuerzburg.de, ORCID 0000-0002-6719-8635]
bibliography: book.bib
params:
  save_globals: FALSE
clean: false
output:
  bookdown::pdf_document2:
    keep_tex: true
---

# Introduction
This document presents graphical results generated as output from analytic workflows written in *R* language and available in [the public GitHub repository - https://github.com/TomVuod/sangchem/tree/supp](https://github.com/TomVuod/sangchem/tree/supp).   
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.align = 'center',
  width = 100,
  echo=FALSE,
  fig.show='hold', 
  fig.keep='all',
  message = FALSE,
  warning = FALSE
#  cache=TRUE
)
```
```{r echo=FALSE}
options("width"=100)
```



<!--chapter:end:/home/user/Documents/Ekologia chemiczna kolonii inicjalnych F. sanguinea/sangchem/inst/supp_materials/index.Rmd-->

# Changes in CHC amount over time

This section contains a series of linear (mixed) models fitted to examine how the proportion of *F. sanguinea* workers in a colony is associated with the amount of CHC, analyzed as a whole or in subsets. Included are summary reports, results of the Shapiro-Wilk test of normality of conditional residuals, as well as diagnostic plots and tests generated with the use of `DHARMa` R package[@R-DHARMa]. The response variable often needed to be transformed to meet model assumptions. Random terms with no variance have been dropped. 

In model specification $(1|\text{random_factor})$ denotes random intercept (separate for each level of the random factor) and $((1|\text{random_factor_1:random_factor_2}))$ denotes random intercepts generated from the combinations of two factors. $\beta_{0}$ and $\epsilon$ denote intercept and error term, respectively. 

```{r}
suppressMessages({
library(sangchem)
library(dplyr)
library(lme4)
library(lmerTest)
library(DHARMa)
})
# load data
data("development_data")
data("body_surface_area")
CHC_normalizer <- get_body_area_normalizer(body_surface_area, development_data)
load_globals()
dev_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==0) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width)) %>% mutate(head_width=head_width*1e-3)
```

## Colony growth and body size

n the analyses below, normalized CHC amounts were used as the response variable. Because body shape in both species deviates from an isometric growth pattern (see Section \@ref(surface-area)), normalizing CHC amounts by the square of head width may introduce bias. To address this, we first examined the relationship between body size (head width) and the proportion of F. sanguinea workers in a colony, and then incorporated the square of head width as an explanatory variable in the linear model.

$$
\text{head_width} =  \beta_0 + \text{sanguinea_proportion} + (1|\text{colony_ID}) + \epsilon,
$$

```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
lm_model <- lmer(head_width ~ sang_prop  + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

## Change in total CHC mass

In this series of models the total CHC mass (normalized by assuming head width of 1.3 mm) was regressed on the proportion of *Formica sanguinea* workers in the colony. 

### Mature *F. sanguinea*

$$
\log(\text{CHC_mass_sanguinea}) =  \beta_0 + \text{sanguinea_proportion} +\newline (1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_sanguinea}$ denotes the total normalized CHC mass on the body of adult *F. sanguinea* worker.

```{r}
set.seed(3453)
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop  +(1|colony) +(1 | colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```
The model shows no significant effect of body size on CHC amount in *F. sanguinea*, and this term will therefore be excluded from subsequent models for this species.

### *F. fusca*

$$
\log(\text{CHC_mass_fusca}) =  \beta_0 + \text{sanguinea_proportion} + (1|\text{colony_ID}) + \epsilon,
$$
where $\text{CHC_mass_fusca}$ denotes the total normalized CHC mass on the body of adult *F. fusca* worker. 

```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop  +(1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### Callow *F. sanguinea*

$$
\log(\text{CHC_mass_callow}) =  \beta_0 + \text{sanguinea_proportion} + \newline(1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_callow}$ denotes the total normalized CHC mass on the body of callow *F. sanguinea* worker.

```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width))

lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```
Similarly to mature *F. sanguinea* ants, in callow workers there is negative but not significant relationship between body size and normalized CHC amount.

### Predictions from the linear model

The linear models explaining the change in CHC amounts in relation to the proportion of *F. sanguinea* workers in a colony were used to predict the normalized CHC amounts of workers of both species in pure (monospecific) colonies. The predictions were obtained by sampling the distribution of the response variable with random effects integrated out. The prediction distributions for both species were matched element-wise, and for each pair of values the CHC mass of *F. sanguinea* was divided by the CHC mass of *F. fusca* to generate the distribution of ratios of the predicted normalized mass. 

```{r}
data("CHC_mass_prediction")
mass_ratios = CHC_mass_prediction$pred_values_sang/CHC_mass_prediction$pred_values_fusca
library(HDInterval)
hdi_interval <- hdi(mass_ratios)
ratio_density <- density(mass_ratios, n=10^3)
plot_data <- data.frame(x = ratio_density$x, 
                        y = ratio_density$y,
                        hdi_mask =ratio_density$x>hdi_interval[1] & ratio_density$x<hdi_interval[2])
```
```{r fig.cap="Ditribution of the ratios (\\textit{F. sanguinea}/\\textit{F. fusca}) of the predicted CHC mass of workers from pure colonies. The shaded area corresponds to the 95% highest density interval."}
ggplot(plot_data, aes(x=x, y=y))+
  geom_line()+
  scale_x_continuous(limits = c(1, max(plot_data$x)))+
  geom_area(aes(x = x, y = y),
            data = subset(plot_data, hdi_mask), fill = "#aaee99")+
  ylab("Density")+
  xlab("Ratio of the normalized CHC amount")+
  theme(panel.border = element_rect(color = "black", fill=NA),
        panel.background = element_rect(color="black", fill = "white"))
```



## Change in CHC characteristic of *F. sanguinea*

In this section the subset of peaks were used to calculate normalized CHC mass. These were peaks identified as *F. sanguinea* markers.

### Mature *F. sanguinea* ants

$$
\log(\text{CHC_mass_mature}) =  \beta_0 + \text{sanguinea_proportion} +  \newline(1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_mature}$ denotes the normalized mass of *F. sanguinea* markers on the body of adult *F. sanguinea* worker.

```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lmer(I(log(normalized_mass+1)) ~ sang_prop +(1 | colony:census_date) + (1|colony), data=model_input)
print(summary(lm_model))
# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### *F. fusca* ants

$$
\sqrt{(\text{CHC_mass_fusca})} =  \beta_0 + \text{sanguinea_proportion} +  \newline(1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_fusca}$ denotes the normalized mass of *F. sanguinea* markers on the body of adult *F. sanguinea* worker.

```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lmer(I(sqrt(normalized_mass)) ~ sang_prop +(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### Callow *F. sanguinea* ants

$$
\sqrt[3]{(\text{CHC_mass_mature})} = \beta_0 + \text{sanguinea_proportion} + \epsilon,
$$
where $\text{CHC_mass_mature}$ denotes the normalized mass of *F. sanguinea* markers on the body of callow *F. sanguinea* worker.

```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width))
# random term dropped because of the model singularity
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lm(I((normalized_mass)^(1/3)) ~ sang_prop, data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
```

## Change in CHC characteristic of *F. fusca*

Similarly to the analysis with the use of markers of *F. sanguinea* workers, the procedure was repeated using the mass of *F. fusca* markers as a response variable. 

### Mature *F. sanguinea* ants

$$
\log(\text{CHC_mass_mature}) =  \beta_0 + \text{sanguinea_proportion} + (1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_mature}$ denotes the normalized mass of *F. fusca* markers on the body of adult *F. sanguinea* worker.


```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop +(1 | colony:census_date) + (1 | colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n=1e4)
plot(simulationOutput)
```

### *F. fusca* ants

$$
\log(\text{CHC_mass_mature}+0.01) =  \beta_0 + \text{sanguinea_proportion} + \newline(1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_mature}$ denotes the normalized mass of *F. fusca* markers on the body of *F. fusca* worker.

```{r}
set.seed(1001)
model_input <- dev_samples[dev_samples$species=="F. fusca",]
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(normalized_mass+0.01)) ~ sang_prop + (1 | colony)+ (1 | colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n=1e4)
plot(simulationOutput)
```

### Callow *F. sanguinea* ants

$$
\log(\text{CHC_mass_mature}) = \beta_0 + \text{sanguinea_proportion} + (1|\text{colony_ID}) + \epsilon,
$$
where $\log(\text{CHC_mass_mature})$ denotes the normalized mass of *F. fusca* markers on the body of callow *F. sanguinea* worker.

```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width))
# random term dropped because of the model singularity
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop+(1 | colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
```

<!--chapter:end:/home/user/Documents/Ekologia chemiczna kolonii inicjalnych F. sanguinea/sangchem/inst/supp_materials/CHCs_over_time.Rmd-->

# Effect of dummy ants

In these experiments, *F. sanguinea* ants released from their cocoon envelopes were maintained in Petri dishes along with glass beads coated with CHC extracted from either *F. fusca* or *F. sanguinea* ants. We examined whether the contact with dummy nestmates influenced the development of CHC profile of young *F. sanguinea* workers. In particular, we aimed to determine there there is evidence for active chemical mimicry in *F. sanguinea*. 

```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(dplyr)
  library(kableExtra)
})

# load data
data("separation_experiment_data")
data("mass_spectra_data")
data("subject_to_donor_mapping")
```

## Distance to the CHC profile of the dummy ants treated with *F. fusca* CHC

We computed the chemical distance to the CHC profile of ants whose CHC were used to cover glass beads serving as dummy ants. In the control variant, we used the ants subjected to clean glass beads. 

```{r}
args <- test_dummy_ant_effect(treatment_id=5, separation_experiment_data, mass_spectra_data, subject_to_donor_mapping,caption="Chemical distance of separated \\textit{F. sanguinea} ants to the CHC profile of \\textit{F. fusca} ants from colonies that served as a source of CHC to caot the glass beads. In control variant, glass bead were left clean.")
args$booktabs=TRUE
```


```{r}
do.call(kableExtra::kbl, args) %>% column_spec(c(3,5), width="2cm") %>% kable_styling()
```
## Distance to the CHC profile of the dummy ants treated with *F. sanguinea* CHC
```{r}
args <- test_dummy_ant_effect(treatment_id=7, separation_experiment_data, mass_spectra_data, subject_to_donor_mapping,caption="Chemical distance of separated \\textit{F. sanguinea} ants to the CHC profile of \\textit{F. sanguinea} ants from colonies that served as a source of CHC to caot the glass beads. In control variant, glass bead were left clean.")
args$booktabs=TRUE
do.call(kableExtra::kbl, args) %>% column_spec(c(3,5), width="2cm") %>% kable_styling()
```



## Fraction of *n*-docosane in the CHC profile

For the control treatment, glass beads were left uncoated with CHC. In the experimental treatments, the CHC used to coat the glass beads were supplemented with *n*-docosane, which served as an internal standard. We examined differences in the proportion of *n*-docosane in the CHC profiles of the tested ants, as these could indicate the acquisition of chemicals from the dummy ants.

### *F. fusca*

```{r}
args <- test_C22_proportion(5, separation_experiment_data, mass_spectra_data,caption="Proportion of \\textit{n}-docosane in CHC extracted from \\textit{F. sanguinea} ants maintained with the glass beads coated with the CHC of \\textit{F. fusca} ants and contaminated with \\textit{n}-docosane. In control variant, glass bead were left clean.")
args$booktabs=TRUE
do.call(kableExtra::kbl, args) %>% column_spec(c(3,5), width="2cm") %>% kable_styling()
```

### *F. sanguinea*

```{r}
args <- test_C22_proportion(7, separation_experiment_data, mass_spectra_data,caption="Proportion of \\textit{n}-docosane in CHC extracted from \\textit{F. sanguinea} ants maintained with the glass beads coated with the CHC of \\textit{F. sanguinea} ants and contaminated with \\textit{n}-docosane. In control variant, glass bead were left clean.")
args$longtable = TRUE
args$booktabs=TRUE
do.call(kableExtra::kbl, args) %>% column_spec(c(3,5), width="2cm") %>% kable_styling()
```

<!--chapter:end:/home/user/Documents/Ekologia chemiczna kolonii inicjalnych F. sanguinea/sangchem/inst/supp_materials/Dummy_ants.Rmd-->

# Body surface area of *F. sanguinea* ants and their slaves {#surface-area}

The areas of planar projection of different body parts were regressed on the head width to obtain a model which was subsequently used to approximate the area for each sample. The regression was performed separately for each species. The following formula was applied:

$$
\hat{A} = \sum_{i \in I}\beta_{0,i}+\beta_{1,i}w+\beta_{2,i}w^2,
$$
where\newline
$\hat{A}$ denotes the estimated body area,\newline
$I$ denotes the set of body parts for which the planar projection area was measured\newline
$w$ denotes head width\newline
$\beta_{j,i}$ are coefficients from the polynomial regression.\newline

These calculations do not reflect the absolute value of body surface area but were converted into scaling factors by using them a divisors of a standard area. In this way, the unit of measurement is canceled out, and absolute values are converted into dimensionless ratios. The standard area was calculated by substituting the $w$ = 1.15 mm into the formula above and using coefficients obtained from measurements of *F. sanguinea*. 

```{r warning=FALSE}
library(sangchem)
library(ggplot2)
data("development_data")
data("separation_experiment_data")
data("body_surface_area")

body_surface_area$species <- development_data$species[match(body_surface_area$ID, development_data$chromatogram_ID)]
body_surface_area$head_width <- development_data$head_width[match(body_surface_area$ID, development_data$chromatogram_ID)]
body_surface_area$head_width <- body_surface_area$head_width*1e-3

```
```{r fig.cap="Relation between head width and area of the planar projections of ant body parts.",fig.height=3}
ggplot(body_surface_area, aes(y=area, x=head_width,color=species))+
  geom_point()+
  stat_smooth(method='lm', formula=y~poly(x,2,raw = TRUE), se=FALSE) +
  facet_wrap(~view)+
  scale_color_manual(values=c("black", "red"))+
  xlab("Head width [mm]")+
  ylab("Area of the planar projection [pixels]")+
  theme_light()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

<!--chapter:end:/home/user/Documents/Ekologia chemiczna kolonii inicjalnych F. sanguinea/sangchem/inst/supp_materials/Body_surface.Rmd-->

