# Changes in CHC amount over time

This section contains a series of linear (mixed) models fitted to examine how the proportion of *F. sanguinea* workers in a colony is associated with the amount of CHC, analyzed as a whole or in subsets. Included are summary reports, results of the Shapiro-Wilk test of normality of conditional residuals, as well as diagnostic plots and tests generated with the use of `DHARMa` R package[@R-DHARMa]. The response variable often needed to be transformed to meet model assumptions. Random terms with no variance have been dropped. 

In model specification $(1|\text{random_factor})$ denotes random intercept (separate for each level of the random factor) and $((1|\text{random_factor_1:random_factor_2}))$ denotes random intercepts generated from the combinations of two factors. $\beta_{0}$ and $\epsilon$ denote intercept and error term, respectively. 

```{r}
suppressMessages({
library(sangchem)
library(dplyr)
library(lme4)
library(lmerTest)
library(DHARMa)
library(ggplot2)
})
# load data
data("development_data")
data("body_surface_area")
CHC_normalizer <- get_body_area_normalizer(body_surface_area, development_data)
load_globals()
dev_samples <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==0) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width)) %>% mutate(head_width=head_width*1e-3)
```

## Colony growth and body size

n the analyses below, normalized CHC amounts were used as the response variable. Because body shape in both species deviates from an isometric growth pattern (see Section \@ref(surface-area)), normalizing CHC amounts by the square of head width may introduce bias. To address this, we first examined the relationship between body size (head width) and the proportion of F. sanguinea workers in a colony, and then incorporated the square of head width as an explanatory variable in the linear model.

$$
\text{head_width} =  \beta_0 + \text{sanguinea_proportion} + (1|\text{colony_ID}) + \epsilon,
$$

```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
lm_model <- lmer(head_width ~ sang_prop  + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

## Change in total CHC mass

In this series of models the total CHC mass (normalized by assuming head width of 1.3 mm) was regressed on the proportion of *Formica sanguinea* workers in the colony. 

### Mature *F. sanguinea*

$$
\log(\text{CHC_mass_sanguinea}) =  \beta_0 + \text{sanguinea_proportion} +\newline (1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_sanguinea}$ denotes the total normalized CHC mass on the body of adult *F. sanguinea* worker.

```{r}
set.seed(3453)
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop  +(1|colony) +(1 | colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```
The model shows no significant effect of body size on CHC amount in *F. sanguinea*, and this term will therefore be excluded from subsequent models for this species.

### *F. fusca*

$$
\log(\text{CHC_mass_fusca}) =  \beta_0 + \text{sanguinea_proportion} + (1|\text{colony_ID}) + \epsilon,
$$
where $\text{CHC_mass_fusca}$ denotes the total normalized CHC mass on the body of adult *F. fusca* worker. 

```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop  +(1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### Callow *F. sanguinea*

$$
\log(\text{CHC_mass_callow}) =  \beta_0 + \text{sanguinea_proportion} + \newline(1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_callow}$ denotes the total normalized CHC mass on the body of callow *F. sanguinea* worker.

```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width))

lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop + (1|colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```
Similarly to mature *F. sanguinea* ants, in callow workers there is negative but not significant relationship between body size and normalized CHC amount.

### Predictions from the linear model

The linear models explaining the change in CHC amounts in relation to the proportion of *F. sanguinea* workers in a colony were used to predict the normalized CHC amounts of workers of both species in pure (monospecific) colonies. The predictions were obtained by sampling the distribution of the response variable with random effects integrated out. The prediction distributions for both species were matched element-wise, and for each pair of values the CHC mass of *F. sanguinea* was divided by the CHC mass of *F. fusca* to generate the distribution of ratios of the predicted normalized mass. 

```{r}
data("CHC_mass_prediction")
mass_ratios = CHC_mass_prediction$pred_values_sang/CHC_mass_prediction$pred_values_fusca
library(HDInterval)
hdi_interval <- hdi(mass_ratios)
ratio_density <- density(mass_ratios, n=10^3)
plot_data <- data.frame(x = ratio_density$x, 
                        y = ratio_density$y,
                        hdi_mask =ratio_density$x>hdi_interval[1] & ratio_density$x<hdi_interval[2])
```
```{r fig.cap="Ditribution of the ratios (\\textit{F. sanguinea}/\\textit{F. fusca}) of the predicted CHC mass of workers from pure colonies. The shaded area corresponds to the 95% highest density interval."}
ggplot(plot_data, aes(x=x, y=y))+
  geom_line()+
  scale_x_continuous(limits = c(1, max(plot_data$x)))+
  geom_area(aes(x = x, y = y),
            data = subset(plot_data, hdi_mask), fill = "#aaee99")+
  ylab("Density")+
  xlab("Ratio of the normalized CHC amount")+
  theme(panel.border = element_rect(color = "black", fill=NA),
        panel.background = element_rect(color="black", fill = "white"))
```



## Change in CHC characteristic of *F. sanguinea*

In this section the subset of peaks were used to calculate normalized CHC mass. These were peaks identified as *F. sanguinea* markers.

### Mature *F. sanguinea* ants

$$
\log(\text{CHC_mass_mature}) =  \beta_0 + \text{sanguinea_proportion} +  \newline(1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_mature}$ denotes the normalized mass of *F. sanguinea* markers on the body of adult *F. sanguinea* worker.

```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lmer(I(log(normalized_mass+1)) ~ sang_prop +(1 | colony:census_date) + (1|colony), data=model_input)
print(summary(lm_model))
# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### *F. fusca* ants

$$
\sqrt{(\text{CHC_mass_fusca})} =  \beta_0 + \text{sanguinea_proportion} +  \newline(1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_fusca}$ denotes the normalized mass of *F. sanguinea* markers on the body of adult *F. sanguinea* worker.

```{r}
model_input <- dev_samples[dev_samples$species=="F. fusca",]
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lmer(I(sqrt(normalized_mass)) ~ sang_prop +(1|colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n = 10^4)
plot(simulationOutput)
```

### Callow *F. sanguinea* ants

$$
\sqrt[3]{(\text{CHC_mass_mature})} = \beta_0 + \text{sanguinea_proportion} + \epsilon,
$$
where $\text{CHC_mass_mature}$ denotes the normalized mass of *F. sanguinea* markers on the body of callow *F. sanguinea* worker.

```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width))
# random term dropped because of the model singularity
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_sang, model_input$chromatogram_ID)
lm_model <- lm(I((normalized_mass)^(1/3)) ~ sang_prop, data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
```

## Change in CHC characteristic of *F. fusca*

Similarly to the analysis with the use of markers of *F. sanguinea* workers, the procedure was repeated using the mass of *F. fusca* markers as a response variable. 

### Mature *F. sanguinea* ants

$$
\log(\text{CHC_mass_mature}) =  \beta_0 + \text{sanguinea_proportion} + (1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_mature}$ denotes the normalized mass of *F. fusca* markers on the body of adult *F. sanguinea* worker.


```{r}
model_input <- dev_samples[dev_samples$species=="F. sanguinea",]
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop +(1 | colony:census_date) + (1 | colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n=1e4)
plot(simulationOutput)
```

### *F. fusca* ants

$$
\log(\text{CHC_mass_mature}+0.01) =  \beta_0 + \text{sanguinea_proportion} + \newline(1|\text{colony_ID}) + (1|\text{colony_ID:sampling_occasion_ID}) + \epsilon,
$$
where $\text{CHC_mass_mature}$ denotes the normalized mass of *F. fusca* markers on the body of *F. fusca* worker.

```{r}
set.seed(1001)
model_input <- dev_samples[dev_samples$species=="F. fusca",]
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(normalized_mass+0.01)) ~ sang_prop + (1 | colony)+ (1 | colony:census_date), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
simulationOutput <- simulateResiduals(fittedModel = lm_model, plot = FALSE, n=1e4)
plot(simulationOutput)
```

### Callow *F. sanguinea* ants

$$
\log(\text{CHC_mass_mature}) = \beta_0 + \text{sanguinea_proportion} + (1|\text{colony_ID}) + \epsilon,
$$
where $\log(\text{CHC_mass_mature})$ denotes the normalized mass of *F. fusca* markers on the body of callow *F. sanguinea* worker.

```{r}
model_input <- filter(development_data, caste=="worker",!is.na(head_width), !remarks %in% c("aggression actor", "aggression target", "queenless colony"), callow==1) %>%
  dplyr::select(chromatogram_ID, species, colony, sang_prop, head_width, mass, census_date) %>%
  mutate(normalized_mass=mass*CHC_normalizer(species, head_width))
# random term dropped because of the model singularity
model_input$normalized_mass <- model_input$normalized_mass*subsample_proportion(peaks_fusca, model_input$chromatogram_ID)
lm_model <- lmer(I(log(normalized_mass)) ~ sang_prop+(1 | colony), data=model_input)
print(summary(lm_model))

# model diagnostics
hist(residuals(lm_model), main="Model conditional residuals", xlab="Residuals")
print(shapiro.test(residuals(lm_model)))
```
