---
title: "Impact of slave on the development of the F. sanguinea CHC profile"
author: "Tomasz Włodarczyk"
date: "4-12-2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r warning=FALSE}
suppressMessages({
  library(sangchem)
})
```

Now we are going to analyze the impact of *F. fusca* slaves on the CHC profile of the callow *F. sanguinea* ants. In the experiment, *F. sanguinea* ants were isolated in pairs before hatching and their CHC profile was analysed at one of the four fixed points of adult life. We want to see whether the callow *F. sanguinea* ants are more similar to slave relatives from free-living colony as compared to *F. fusca*  ants unrelated at all.
```{r}
# Load data
data("separation_experiment_data")
data("slave_source_colony")
data("colony_to_chrID")

# Exclude colonies for which we don't have free-living F. fusca CHC profile
separation_data <- separation_experiment_data[!is.na(colony_to_chrID[slave_source_colony[separation_experiment_data$colony]]),]

# See the lifespan of the workers depending on the treatment
plot_data <- separation_data[separation_data$treatment_id %in% 1:4,]
plot_data$treatment <- factor(plot_data$treatment, levels = c("1-3 days", "8-10 days", "17-20 days", "35-40 days"))
boxplot(plot_data$mean_delta~plot_data$treatment, main = "Adult lifespan", xlab = "Treatment", ylab = "Separation period after hatching (avegared for a pair of individuals)")

# Calculate emprical p-value for the treatment 1
res_1 <- distances_permutation_test(treatment_id = 1)
print(res_1$p.val)

# Calculate emprical p-value for the treatment 2
res_2 <- distances_permutation_test(treatment_id = 2)
print(res_2$p.val)

# Calculate emprical p-value for the treatment 3
res_3 <- distances_permutation_test(treatment_id = 3)
print(res_3$p.val)

# Calculate emprical p-value for the treatment 3
res_4 <- distances_permutation_test(treatment_id = 4)
print(res_4$p.val)
```