---
title: "Finding peaks discriminating species"
author: "Tomasz Włodarczyk"
date: "24-11-2023"
output: html_document
params:
  save_globals: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Ants in mixed colonies mixed CHCs to produce hybrid chemical identity. Here we develop the method to assign each sample a number indicating how much it is characteristic of one or the other species.
```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(stringi)
  library(dplyr)
  library(mixOmics)
})

# load data from the the earlier study
data("field_colonies")

# select samples from F. sanguinea ants and free-living F. fusca ants encoded in chromatogram_ID
fusca_samples <- as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID), "fus[0-9].?-[0-9]\\([1-2]\\)")])
sanguinea_samples<-as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID),"sangFs[0-9].?\\([0-9]\\)")])

# remove outliers
sanguinea_samples <- setdiff(sanguinea_samples,c("sangFs2(1)","sangFs5(2)","sangFs26(1)","sangFs26(2)"))
fusca_samples <- setdiff(fusca_samples,c("fus9-2(1)", "fus26-3(2)"))
discrimination_data <- filter(field_colonies,chromatogram_ID %in% c(fusca_samples,sanguinea_samples)) 
peak_prop_unfiltered <- peak_proportions_table(discrimination_data)
# feature selection
peak_prop <- freq_abun_QC(peak_prop_unfiltered)
# re-normalize rows to 1
peak_prop <- peak_prop/rowSums(peak_prop)
```

The main idea is to fit a sPLS_DA model to the training set (sampled from field colonies) and use it predict the explanatory variable (species) in the test data (samples from experimental colonies). However the sPLS_DA model run on the raw data will yield the classificator which used only few variables (peaks) due to the well separation of samples of both species. The unused variables might also be informative in the species discrimination task. We want to use as much data as it is possible to make robust predictions and thus we will use principal component analysis output. The same rotation matrix will be then used to transform test data. 

```{r}
PCA_species_discr <- prcomp(peak_prop, scale=FALSE)
```

We also need to deifne our reponse variable to be explained by discrimination model.
```{r}
#define groups to be compared
Y <- rownames(PCA_species_discr$x) %in% sanguinea_samples
```

Now that we have prepared the input data we can run the model. The following workflow is based on the tutorial available at the official [`mixOmics` package web site](https://mixomics.org/mixmc/hmp-bodysites-case-study/).
```{r}
# run the main model
species_discrimination_PC.plsda <- plsda(PCA_species_discr$x, Y, ncomp=10, scale=FALSE)
plotIndiv(species_discrimination_PC.plsda,group=Y,
          legend=TRUE,ellipse=TRUE,comp=1:2)

# determine the optimal number of components used in the model
spec_discr_PC.plsda.perf<-perf(species_discrimination_PC.plsda, nrepeat = 10,folds=5,
                            progressBar = FALSE,validation = "Mfold",
                            auc=TRUE,scale=FALSE)

plot(spec_discr_PC.plsda.perf, sd = TRUE, overlay = 'measure')
```

The diagnostic plots shows that using one compoment is enough to accurately discriminate between species.
```{r}
selected_n <- 3

# dtermine the number of the original variables to be kept in each component (actually
# here we have only one component)
spec_discr_PC.splsda <- tune.splsda(PCA_species_discr$x,Y,test.keepX = c(1:60),ncomp = selected_n,
                                 validation="Mfold", progressBar=FALSE, dist="max.dist", 
                                 nrepeat=5, cpus=2, folds=5, measure="BER",
                                 scale=FALSE)

# final model
species_prediction_model <- splsda(PCA_species_discr$x, Y, ncomp=selected_n,
                           keepX =spec_discr_PC.splsda$choice.keepX,
                           scale=FALSE)

# visuzalize results for the first two commponents
plotIndiv(species_prediction_model, group=Y, legend=TRUE, ellipse=TRUE, comp=1:2)

```
```{r eval=params$save_globals, echo=params$save_globals}
globals <- readRDS(system.file("globals.rds", package="sangchem"))
globals[["species_prediction_model"]] <- species_prediction_model
globals[["PCA_species_discr"]] <- PCA_species_discr
saveRDS(globals, system.file("globals.rds", package="sangchem"))
```

```{r}
print(sessionInfo())
```
