---
title: "Finding peaks discriminating species"
author: "Tomasz Włodarczyk"
date: "24-11-2023"
output: html_document
params:
  save_globals: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r warning=FALSE}
suppressMessages({
  library(sangchem)
  library(stringi)
  library(dplyr)
  library(mixOmics)
})

# load data from the the earlier study
data("field_colonies")

# select samples from F. sanguinea ants and free-living F. fusca ants encoded in chromatogram_ID
fusca_samples <- as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID), "fus[0-9].?-[0-9]\\([1-2]\\)")])
sanguinea_samples<-as.character(unique(field_colonies$chromatogram_ID)[stri_detect_regex(unique(field_colonies$chromatogram_ID),"sangFs[0-9].?\\([0-9]\\)")])

# remove outliers
sanguinea_samples <- setdiff(sanguinea_samples,c("sangFs2(1)","sangFs5(2)","sangFs26(1)","sangFs26(2)"))
fusca_samples <- setdiff(fusca_samples,c("fus9-2(1)", "fus26-3(2)"))
discrimination_data <- filter(field_colonies,chromatogram_ID %in% c(fusca_samples,sanguinea_samples)) 
peak_prop_unfiltered <- peak_proportions_table(discrimination_data)
# feature selection
peak_prop <- freq_abun_QC(peak_prop_unfiltered)
# re-normalize rows to 1
peak_prop <- peak_prop/rowSums(peak_prop)
# replace zeros with a small number before transformation
peak_prop[peak_prop==0] <- 10^-16
# apply central log ratio transformation
preak_prop <- apply(peak_prop, 1, clr_transformation)


#define groups to be compared
Y <- rownames(peak_prop) %in% sanguinea_samples
```

The following workflow is based on the tutorial available at the official [`mixOmics` package web site](https://mixomics.org/mixmc/hmp-bodysites-case-study/).
```{r}
# run the main model
species_discrimination.plsda <- plsda(peak_prop, Y, ncomp=10, logratio = "CLR",scale=FALSE)
plotIndiv(species_discrimination.plsda,group=Y,
          legend=TRUE,ellipse=TRUE,comp=1:2)

# determine the optimal number of components used in the model
spec_discr.plsda.perf<-perf(species_discrimination.plsda, nrepeat = 10,folds=5,
                            progressBar = FALSE,validation = "Mfold",
                            auc=TRUE,scale=FALSE)

plot(spec_discr.plsda.perf, sd = TRUE, overlay = 'measure')
```

The diagnostic plots shows that using one compoment is enough to accurately discriminate between species.
```{r}
selected_n <- 1

# dtermine the number of the original variables to be kept in each component (actually
# here we have only one component)
spec_discr.splsda <- tune.splsda(peak_prop,Y,test.keepX = c(1:60),ncomp = selected_n, validation="Mfold", progressBar=FALSE, dist="max.dist", nrepeat=5, cpus=2, logratio = "CLR", folds=5, measure="BER", scale=FALSE)

# final model
spec_discr.plsda_2 <- splsda(peak_prop,Y,ncomp=selected_n,
                           keepX =spec_discr.splsda$choice.keepX,logratio = "CLR",
                           scale=FALSE)

```

Now we want to correlate relative abundances of CHCs with the species predicted by the model for each sample.
```{r message=FALSE}
# quantify predicted species (F. sanguinea closer to one and F. fusca closer to 0)
predicted_species <- predict(spec_discr.plsda_2, peak_prop)$predict[,2,1]

library(Hmisc) 
cor_p_val<-rcorr(cbind(peak_prop, predicted_species))
cor_p_val$P[cor_p_val$r==1]<-0
#significance threshold = 0.05
significant_correlations <- cor_p_val$r["predicted_species",][cor_p_val$P["predicted_species",]<0.05]
peaks_sang <- names(significant_correlations)[significant_correlations>0]
peaks_sang <- as.numeric((setdiff(peaks_sang,"predicted_species")))
peaks_fusca <- as.numeric(names(significant_correlations)[significant_correlations<0])
print(peaks_sang)
print(peaks_fusca)
```

The objects `peaks_sang` and `peaks_fusca` are stored in the package and can be collectively loaded into the global environment along with the other objects by running `load_globals()`.
```{r eval=params$save_globals, echo=params$save_globals}
globals <- readRDS(system.file("globals.rds", package="sangchem"))
globals[["peak_sang"]] <- peaks_sang
globals[["peak_fusca"]] <- peaks_fusca
saveRDS(globals, system.file("globals.rds", package="sangchem"))
```
```{r}
print(sessionInfo())
```
